---
title: "Session 3: Advanced Epidemic Curves & Spatial Mapping"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

This session focuses on the visual and spatial dimensions of field epidemiology. Students will progress from exploratory continuous plotting to constructing advanced, publication-ready temporal and geographical models used to brief ministry leaders during an active emergency.

| Aim | Tool / function |
|--------------------------|----------------------------------------------|
| 1\. Construct stratified temporal plots | **ggplot2::geom_histogram()** & **geom_col()** |
| 2\. Smooth operational artifacts | **slider::slide_index_dbl()** rolling curves |
| 3\. Manage geospatial data vectors | **sf::st_as_sf()** & **sf::st_transform()** |
| 4\. Defeat the "Choropleth Trap" | Population Denominator Standardization |
| 5\. Deploy dynamic web visualizations | **plotly::ggplotly()** interactive widgets |

# Packages used for this session

```{r}
pacman::p_load(
  # ── Core Infrastructure & Data I/O ─────────────────────────────────────────
  here,         # Relative file pathways to prevent broken working directory links
  rio,          # Swiss-Army knife for importing/exporting data (xlsx, csv, rds)
  tidyverse,    # Core ecosystem (ggplot2, dplyr, tidyr) for data science & viz
  
  # ── Data Cleaning & Standardisation ────────────────────────────────────────
  janitor,      # Clean column names, find duplicates, and build quick tabyls
  lubridate,    # Parsing and executing calculations on epidemic date vectors
  
  # ── Spatial Mapping & Geographics ──────────────────────────────────────────
  sf,           # Handling Simple Features geospatial vector data and polygons
  scales,       # Internal label formatting formatting (e.g., shorthand axis percentages)
  
  # ── Advanced Visualization ─────────────────────────────────────────────────
  plotly,       # Rendering static ggplot matrices into interactive HTML widgets
  slider,       # Calculating sliding windows for trailing rolling averages
  RColorBrewer  # Accessing color palettes designed for categorical features
)
```

# Load data

We import the cleaned cholera line-list generated during our data-cleaning modules. We verify that all date vectors are explicitly recognized by the R environment.

```{r}
cholera_linelist <- import(here("processed_data", "cholera_linelist_clean.xlsx")) %>%
  mutate(
    onset_date = as.Date(onset_date),
    # Pre-clean coordinates to protect mapping modules
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)
  )
```

# Epidemic Curves: Temporal Dynamics of Outbreaks

An Epidemic Curve (Epi-curve) is a visual histogram of an outbreak that plots the number of emergent cases along the Y-axis against their specific date of symptom onset along the X-axis.

An epi-curve provides essential visual clues about the outbreak's magnitude, outliers, time trends, and primary mode of transmission: Point Source:

- A common source exposure where the transmission window is brief and all cases cluster within a single incubation period.

- Continuous Common Source: Exposure is prolonged or persistent, causing cases to rise gradually and form a plateau rather than a sharp peak.

- Intermittent Common Source: Exposure occurs sporadically, producing irregular, jagged peaks reflecting the timing of the vehicle's availability.

- Propagated Outbreak: Spread person-to-person. These epidemics display progressively taller, compounding waves spaced roughly one incubation period apart.

## Basic Data Visualization: Understanding Age Distribution

Before plotting dates, we evaluate the distribution of underlying continuous vectors like age. This alerts us to missing demographic fields, recording errors, or age skews.

```{r}
ggplot(data = cholera_linelist, mapping = aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "grey50", color = "white") +
  labs(title = "Exploratory Age Distribution Matrix", x = "Age (Years)", y = "Count") +
  theme_minimal()
```

## Basic Epidemic Curve (Daily Intervals)

Setting binwidth = 1 isolated day-by-day case incidences. This granular view is vital during early-phase cluster evaluations to pinpoint exact indexing windows.

```{r}
ggplot(data = cholera_linelist, mapping = aes(x = onset_date)) +
  geom_histogram(binwidth = 1, fill = "darkred", color = "black", linewidth = 0.2) +
  scale_x_date(date_breaks = "1 week", labels = label_date_short()) +
  theme_minimal() +
  labs(title = "Daily Epidemic Curve of Cholera Outbreak", 
       x = "Date of Symptom Onset", 
       y = "Daily Case Incidences")
```

## Stratified Epidemic Curve (Weekly Aggregations)

Extended outbreaks can generate excessive daily variance ("noise") due to reporting delays or weekend closure artifacts. Setting binwidth = 7 pools records into standard 7-day epidemiological blocks. Adding a fill parameter reveals geographic propagation patterns across states over time.

```{r}
ggplot(data = cholera_linelist, mapping = aes(x = onset_date, fill = state_name)) +
  geom_histogram(binwidth = 7, position = "stack", color = "white", linewidth = 0.2) +
  scale_fill_viridis_d(option = "plasma") +
  #scale_fill_brewer(palette = "Dark2") +
  #scale_fill_viridis_d(option = "rocket") +
  scale_x_date(date_breaks = "2 weeks", labels = label_date_short()) +
  labs(title = "Weekly Cholera Cases Stratified by State", 
       x = "Date of Symptom Onset (7-day Bins)", 
       y = "Number of Cases", 
       fill = "Reporting State") +
  theme_minimal()
```

## Categorical Cross-Sectioning: Epi-Weeks by Demographic Group

Instead of relying solely on geom_histogram, we can explicitly calculate standard epidemiological weeks using lubridate and plot them via geom_col(). Here, we track age-group representation across time.

```{r}
cholera_linelist %>%
  filter(!is.na(onset_date)) %>%
  mutate(
    epiweek = floor_date(onset_date, unit = "week", week_start = 1),
    age_cat = cut(age,
                  breaks = c(0, 5, 15, 30, 45, 60, Inf),
                  right  = FALSE,
                  labels = c("0-4", "5-14", "15-29", "30-44", "45-59", "60+"))
  ) %>%
  count(epiweek, age_cat) %>%
  ggplot(aes(x = epiweek, y = n, fill = age_cat)) +
  geom_col(color = "white", linewidth = 0.1) +
  scale_fill_brewer(palette = "Set3", name = "Age Group") +
  scale_x_date(date_breaks = "2 weeks", labels = label_date_short()) +
  labs(title = "Weekly Cholera Distribution by Age Category, 2025",
       x = "Epi-Week (Start Date)", y = "Confirmed Cases (n)") +
  theme_minimal()
```

## Facetted & Segmented Outbreak Overviews

To evaluate competing exposure hazards, we apply facet_wrap(). This creates parallel timelines for each geographic territory, allowing us to see if specific water sources drive distinct outbreak waves in different regions.

```{r}
ggplot(data = cholera_linelist, mapping = aes(x = onset_date, fill = water_source)) +
  geom_histogram(binwidth = 7, position = "stack", color = "white", linewidth = 0.1) +
  scale_fill_brewer(palette = "Set1") + 
  #scale_fill_viridis_d(option = "plasma") +  
  #scale_fill_brewer(palette = "Dark2") +
  #scale_fill_viridis_d(option = "rocket") + 
  labs(title = "Cholera Progression by Water Source across Reporting States", 
       x = "Date of Symptom Onset (7-Day Windows)", 
       y = "Case Counts", 
       fill = "Water Source") +
  theme_minimal() +
  facet_wrap(~ state_name, scales = "free_y") # Adjusts panel scales for regional scale variations
```

## Localised Deep Dive: Focus on Red Sea State

Fianlly, we can now specfiy a single state within our master dataset to create a focused epi-curve. For example, Red Sea State has the highest case counts and a distinct temporal pattern. By filtering our data frame to include only records from Red Sea State, we can create a more detailed epi-curve that highlights the outbreak's progression and clinical outcomes within this specific region.

**Note, that you can always change the varaible mapped to fill to stratify by a different variable, such as patient outcome. This allows us to see not only the number of cases over time but also how the severity of cases may have evolved during the outbreak.**

```{r}
ggplot(data = cholera_linelist %>% filter(state_name == "Red Sea"), mapping = aes(x = onset_date, fill = outcome)) +
  geom_histogram(binwidth = 7, position = "stack", color = "white") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_fill_viridis_d(option = "rocket") +
  labs(title = "Outbreak Profile and Clinical Outcomes within Red Sea State", x = "Date of Symptom Onset", y = "Case Count", fill = "Patient Outcome") +
  theme_minimal()
```

# Geographic Mapping: Visualizing Spatial Spread

## Simple map with case counts by state

To visualize the spatial distribution of cases, we can create a geographic map. This requires geospatial data for the regions of interest, which can be obtained from sources like Natural Earth or GADM. We will use the `sf` package to handle spatial data and `ggplot2` for plotting.

Here is an example of a chloropleth map showing the number of cases per state in Sudan. The code includes steps to ensure that the state names in the line list match those in the spatial data, and it aggregates any unrecognized states into an "Other/Unknown" category.

```{r}
library(dplyr)
library(tidyr)   # for replace_na()
library(sf)

# ── 1.  Vector of valid state names (from the line list) ───────────────
valid_states <- unique(cholera_linelist_clean$state_name)
valid_states
#>  [1] "Blue Nile"     "Gezira"        "Kassala"       "Khartoum"
#>  [5] "North Darfur"  "Other/Unknown" "Red Sea"       "South Darfur"
#>  [9] "White Nile"

# ── 2.  Recode polygons: any name NOT in that vector ⇒ "Other/Unknown" ─
sudan_adm1 <- sudan_adm1 %>%           # your sf object with a 'woe_name' column
  mutate(
    state_name = if_else(
      woe_name %in% valid_states,
      woe_name,              # keep the recognised name
      "Other/Unknown"        # collapse everything else
    )
  )

# ── 3.  Aggregate polygons now labelled "Other/Unknown"  ───────────────
#       (optional but tidy; avoids overlapping geometries)
sudan_adm1_agg <- sudan_adm1 %>%
  group_by(state_name) %>%             # 9 known + 1 "Other/Unknown"
  summarise(geometry = st_union(geometry), .groups = "drop")

# ── 4.  Re-count cases in the line list (collapse unknowns too) ────────
cholera_counts <- cholera_linelist_clean %>%
  mutate(
    state_name = if_else(
      state_name %in% valid_states,
      state_name,
      "Other/Unknown"
    )
  ) %>%
  count(state_name, name = "n_cases")

# ── 5.  Join counts onto polygons (names now align)  ───────────────────
map_df <- sudan_adm1_agg %>%
  left_join(cholera_counts, by = "state_name") %>%
  replace_na(list(n_cases = 0))

# ── 6.  Plot ───────────────────────────────────────────────────────────
ggplot(map_df) +
  geom_sf(aes(fill = n_cases), colour = "grey40", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Cholera Cases per State, Sudan 2025",
       fill  = "Cases") +
  theme_minimal()
```

## Defeating the "Choropleth Trap": Standardized Incidence Mapping

A common pitfall in spatial epidemiology is plotting raw case counts across geographical zones. This approach makes large, densely populated administrative regions look like intense transmission epicenters simply because they have more residents.

To evaluate true biological risk, we introduce a lookup data frame containing official baseline population projection metrics, allowing us to map Incidence Rates per 100,000 population.

```{r}
library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)

# 1. Comprehensive 18-State Master Population Reference
state_pops <- tibble(
  state_name = c(
    "Khartoum", "South Darfur", "Gezira", "North Kordofan", 
    "Kassala", "Central Darfur", "White Nile", "North Darfur", 
    "Al Qadarif", "South Kordofan", "Sennar", "West Darfur", 
    "River Nile", "Red Sea", "West Kordofan", "East Darfur", 
    "Blue Nile", "Northern"
  ),
  population = c(
    7993900, 5353025, 5096920, 3174029, 
    2519071, 2499000, 2493880, 2304950, 
    2208385, 2107623, 1918692, 1775945, 
    1511442, 1482053, 1724897, 1587200, 
    1108391,  936255
  )
)

# 2. Harmonize Shapefile Strings on the fly
sudan_adm1_clean <- sudan_adm1 %>%
  mutate(
    state_name = case_when(
      woe_name == "Southern Darfur" ~ "South Darfur",
      woe_name == "North Kordufan"  ~ "North Kordofan",
      woe_name == "Gedarif"         ~ "Al Qadarif",
      woe_name == "Eastern Darfur"  ~ "East Darfur",
      woe_name == "Western Darfur"  ~ "West Darfur",
      TRUE                          ~ woe_name
    )
  )

# 3. Aggregate Line List Counts (Handling real missingness cleanly)
cholera_counts <- cholera_linelist_clean %>%
  mutate(
    state_name = case_when(
      state_name == "South Darfur"   ~ "South Darfur",
      is.na(state_name)              ~ "Unspecified Location",
      state_name == "Other/Unknown"  ~ "Unspecified Location",
      TRUE                          ~ state_name
    )
  ) %>%
  count(state_name, name = "n_cases")

# 4. The Complete Left-Join (Preserves ALL 18 spatial polygons)
map_df <- sudan_adm1_clean %>%
  left_join(cholera_counts, by = "state_name") %>%
  left_join(state_pops, by = "state_name") %>%
  mutate(
    # True structural zeros: If a state isn't in the linelist, it has 0 cases
    n_cases = replace_na(n_cases, 0),
    # Calculate stable incidence rates across the entire country
    cases_per_100k = (n_cases / population) * 100000
  )

# 5. Plot the complete, non-broken national map
ggplot(map_df) +
  geom_sf(aes(fill = cases_per_100k), colour = "grey40", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", direction = -1, na.value = "grey70") +
  labs(
    title = "National Cholera Incidence Matrix, Sudan 2025",
    subtitle = "Population-standardized risk maps derived from unaltered field line-lists",
    fill  = "Cases per 100k"
  ) +
  theme_minimal()
```

## Granular Dot Density Point Mapping

When raw GPS coordinates are available, point mapping bypasses arbitrary political boundaries to surface actual geographical transmission clusters (e.g., cases concentrating along river vectors or water pipeline pathways).

A map like this can be created by plotting the latitude and longitude of each case as points on top of the state-level polygons. This allows us to see not only the overall distribution of cases by state but also where within those states the cases are occurring. This can be useful to identify hotspots within states and to understand the spatial distribution of cases at a finer scale.

**The only way to get this granualr is if you data contains latitiude and longtitude information for each case.**

```{r}
# Coerce coordinates to spatial points object using WGS84 CRS (4326)
cholera_pts <- cholera_linelist %>%
  # 1. Look at your column names and map them safely here.
  #    Change 'longitutde' or 'lng' to whatever is actually in your spreadsheet.
  rename(
    longitude = longitutde, # <-- Fix if it's actually "long", "Longitude", etc.
    latitude  = latitude   # <-- Fix if it's actually "lat", "Latitude", etc.
  ) %>% 
  
  # 2. Filter out missing coordinates
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  
  # 3. This will now run flawlessly because the names are guaranteed to match
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

# Ensure the boundary maps utilize matching coordinate projections
sudan_adm1_crs <- st_transform(sudan_adm1, 4326)

ggplot() +
  geom_sf(data = sudan_adm1_crs, fill = "grey90", colour = "grey75", linewidth = 0.2) +
  geom_sf(data = cholera_pts, alpha = 0.4, size = 1.2, colour = "darkred") +
  coord_sf(xlim = c(21, 39), ylim = c(9, 22)) +
  labs(title = "Spatial Point Distribution of Confirmed Cholera Cases",
       subtitle = "Sudan Field Surveillance Log, 2025",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

## Interactive Web Mapping via Plotly

Static layout files are excellent for situation reports, but teams on the ground need to zoom and explore hotspots interactively. We use ggplotly with a custom text attribute to convert static charts into interactive, hover-responsive HTML widgets.

```{r}
library(ggplot2)
library(plotly) 

# 1. Build the static ggplot object with color stratification
sf_point_map <- ggplot() +
  # Background layer of Sudan states
  geom_sf(data = sudan_adm1, fill = "grey95", colour = "grey80", linewidth = 0.3) +
  
  # Stratified case points layer
  geom_point(
    data = cholera_pts,
    aes(
      x = longitude, 
      y = latitude, 
      colour = outcome, # CRITICAL: Stratifies points by outcome color-coding
      text = paste("Age:", age, "<br>Outcome:", outcome)
    ),
    alpha = 0.6, 
    size = 1.5
  ) +
  
  # Zoom window parameters for Sudan boundaries
  coord_sf(xlim = c(21, 39), ylim = c(9, 22)) +
  
  # Access a clear color palette for categorical outcomes
  scale_colour_brewer(palette = "Set1") +
  
  labs(
    title = "Spatial Distribution of Cholera Cases by Clinical Outcome",
    x = "Longitude", 
    y = "Latitude",
    colour = "Clinical Outcome"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. Convert to an interactive plotly widget
# tooltip = "text" ensures only your clean custom text string appears on hover
ggplotly(sf_point_map, tooltip = "text")
```

# Group Exercise

**Time allowed:** 45-60 minutes

Working in your assigned groups, you will apply these visual and spatial principles to your active team dataset: `dengue_linelist_2025_sudan.xlsx`.

### Your Objectives

- **Epidemic Curve Construction:** Open your group workspace and generate a baseline daily epidemic curve for the Dengue data. Then, pool the records into 7-day epidemiological blocks.

- **Stratified Facet Analysis:** Facet your weekly epidemic curve by state name, mapping `serotype` to the fill aesthetic.

- **Denominator-Standardized Mapping:** Load the provided Sudan administrative boundary maps. Match your states, join the regional baseline population metrics from today's lesson, and construct a standardized Dengue incidence map per 100,000 population. Contrast this with a raw count map.

- **Interactive Dashboard Component:** Convert your point-source coordinate distribution data into an interactive `plotly` map. Ensure that when users hover over case coordinates, it clearly reports the patient's age, clinical outcome, and mosquito netting use status.

Open `student_code / Session 3: Epi-Curves and Maps_Exercise.qmd` to begin. Be prepared to present your trendlines and standardized maps to the class during our wrap-up discussion!

# Resources

- [**The Epidemiologist R Handbook**](https://epirhandbook.com/en/) The definitive, open-access reference manual for applied public health and field epidemiology in R. Excellent for step-by-step guidance on outbreak data cleaning, epi-curves, and contact tracing pipelines.

- [**NCIPH FOCUS: Understanding and Creating Epidemic Curves**](https://nciph.sph.unc.edu/focus/vol1/issue5/1-5EpiCurves_issue.pdf) A foundational guide from the North Carolina Center for Public Health Preparedness. Essential reading for understanding how incubation periods, transmission modes, and data intervals dictate the shape of an epi-curve.

- [**R Programming for Data Science**](https://bookdown.org/rdpeng/rprogdatascience/) Roger D. Peng’s deep dive into the foundational mechanics of the R language, data types, scoping rules, and functional programming configurations.

- [**Efficient R Programming**](https://csgillespie.github.io/efficientR/) A masterclass by Colin Gillespie and Robin Lovelace on optimizing R code performance. Great for learning how to handle large surveillance datasets, memory profiling, and pipeline acceleration.

- [**Hands-On Programming with R (HOPR)**](https://rstudio-education.github.io/hopr/) Garrett Grolemund’s user-friendly introduction to data science basics. Highly recommended for absolute beginners in the class who need practice with basic data structures, vectors, and loops.

- [**R for Data Science (R4DS)**](https://r4ds.had.co.nz/) The industry-standard text by Hadley Wickham and Garrett Grolemund focusing on the `tidyverse` ecosystem. This is the core reference for data transformation with `dplyr`, reshaping with `tidyr`, and programmatic plotting with `ggplot2`.

- [**Shiny by Posit** ](https://shiny.posit.co)– The official Shiny homepage, with tutorials, galleries, and documentation for building interactive web applications in R.