---
title: "Session 2: Epidemiologic Measures, Tables, Plots & Maps"
author: "Max Wegener"
format: html
editor: visual
---

# Overview  

This session extends the data-cleaning skills from Session 1 into basic outbreak epidemiology and spatial visualisation.  
Skills you will practise:

| Aim | Tool / function |
|-----|-----------------|
| 1. Produce clean descriptive tables | **janitor::tabyl()**, **gtsummary**, **dplyr::count()** |
| 2. Compute measures of association | 2×2 tables ➜ **epitools::oddsratio()** & **epiR::epi.2by2()** |
| 3. Calculate attack rates | **dplyr** pipelines |
| 4. Draw an epidemic curve | **ggplot2** |
| 5. Map state-level incidence | **sf**, **ggplot2**, **scale_fill_viridis_c** |

All code is fully reproducible: press **Render** to rebuild the HTML after any change.

# Packages used for this session
```{r}
pacman::p_load(
  here,        # file paths
  rio,         # import/export
  tidyverse,   # core data wrangling + ggplot2
  janitor,     # tabyl()
  skimr,       # quick inspection
  gtsummary,   # publication-ready tables
  epitools,    # odds ratio
  epiR,        # risk ratio, attack rate helpers
  lubridate,   # date handling
  sf,          # simple features for GIS
  viridis,     # colour scales
  gt,
  janitor,
  epitools
)
```

# Load the cleaned cholera line-list

```{r}
cholera_linelist <- import(here("processed_data", "cholera_linelist_clean.xlsx"))
```

# Descriptive tables

## Simple frequency table

```{r}
cholera_linelist %>% 
  tabyl(state_name, show_missing_levels = TRUE) %>% 
  adorn_pct_formatting()
```

```{r}
cholera_linelist %>% 
  tabyl(sex, outcome)
```

```{r}
cholera_linelist %>% 
  tabyl(sex, outcome) %>% 
        adorn_totals("row") %>%        # optional Total row
        adorn_percentages("row") %>% 
        adorn_pct_formatting(digits = 1) %>% 
        adorn_ns(position = "front") 
```

# Simple statistical tests

## t-test

## Chi-squared test

## Summary statistics

## odds ratio

## risk ratio

## attack rate

# Epidemic Curves

# Maps
















# Measures of association

Assume we want to know whether drinking surface water (vs. other sources) is associated with death.

```{r}

```
