---
title: "Session 2: Summaries and Measures"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

Building on the data-cleaning techniques from Session 1, this session moves into basic descriptive epidemiology: frequency tables, summary measures, and introductory statistical tests. Every code chunk is annotated with inline comments (lines beginning with `#`) that briefly explain what each command does, letting you follow the workflow step by step.

## Exercises

Hands-on practice breaks are built into the lesson. The exercise file for this session is located in `student_code/` and is named:\
“Session 2 - Summaries and Measures_Exercise.qmd”

## Answer Key

Solutions are in the same folder. The answer sheet for this session is: “ANSWERS_Session 2 - Summaries and Measures_Exercise.qmd”

# Packages used for this session

```{r}
pacman::p_load(
  # ── Core Infrastructure & Data I/O ─────────────────────────────────────────
  here,         # Relative file pathways to prevent broken working directory links
  rio,          # Swiss-Army knife for importing/exporting data (xlsx, csv, rds)
  tidyverse,    # Core ecosystem (ggplot2, dplyr, tidyr) for data science & viz
  
  # ── Data Cleaning & Standardisation ────────────────────────────────────────
  janitor,      # Clean column names, find duplicates, and build quick tabyls
  lubridate,    # Parsing and executing calculations on epidemic date vectors

  # ── Field Epidemiology & Summaries ─────────────────────────────────────────
  gtsummary,    # Publication-ready baseline descriptive tables (Table 1)
  rstatix,      # Pipe-friendly framework for running classical hypothesis tests
    
  # ── Spatial Mapping & Geographics ──────────────────────────────────────────
  sf,           # Handling Simple Features geospatial vector data and polygons

  # ── Advanced Visualization ─────────────────────────────────────────────────
  RColorBrewer  # Accessing color palettes designed for categorical features
)
```

# Load exercise data

We will load the cleaned cholera line-list that we created in Session 1.

**Quick Note:**

After re-importing the cleaned cholera line-list for Session 2, force `onset_date` back to class `Date`:

```{r}
# Packages: rio::import(), dplyr::mutate()

cholera_linelist <- import(here("processed_data", "cholera_linelist_clean.xlsx")) %>%
  mutate(onset_date = as.Date(onset_date)) # convert the onset_date column from character/factor to Date class
```

**Why?**\
• `rio::import()` can read Excel dates as character strings or Excel serial numbers.\
• Many time-series tasks—sorting chronologically, calculating intervals, grouping by week, plotting with `ggplot2` date scales—require the `Date` class.\
• Converting immediately with `as.Date()` prevents subtle errors later in epi-curve construction and other temporal analyses.

# Descriptive tables for categorical variables using the tbyl within the janitor package

Descriptive tables are the natural starting point for exploring a new dataset. Before moving on to complex models, it’s helpful to review simple counts, percentages, and missing-value patterns for key variables such as sex, region, and outcome.

In this session we’ll use two helpful tools:

- `tabyl()` from the `janitor` package – quickly produces tidy, easy-to-read frequency tables.

- `get_summary_stats()` from the `rstatix` package - returns common summary statistics (n, mean, SD, median, IQR, etc.) for numeric variables and groups you specify.

Together these functions let you see the shape of your data at a glance and spot any issues that need attention before deeper analysis.

Additionally, the `janitor` package comes with several adorn functions that specifically work with `tabyl()` that enhance tables with additional formatting and information. The different adorn functions include:

1.  adorn_totals() - MUST come first (after tbyl()\] since it needs raw numbers

2.  adorn_percentages() - converts to proportions

3.  adorn_rounding() - if needed, to round proportions

4.  adorn_pct_formatting() - converts to percents and formats

5.  adorn_ns() - adds the raw counts

6.  adorn_title() - adds titles last -\> row_names=" "; col_name=" "; placement="combined"; placement="top"

## Simple frequency table

A simple single-variable frequency table allows us to identify case distributions by geographical area and flag structural formatting errors or empty levels.

```{r}
# Packages: janitor

cholera_linelist %>%          # start with the cleaned line-list data frame
  tabyl(state_name) %>%        # make a frequency table of the `state_name` column
  adorn_pct_formatting()       # add and format percentage columns for readability
```

## Two-way cross-tabulation

Cross-tabulations evaluate the raw intersections of two categorical parameters (e.g., checking outcomes stratified by sex).

```{r}
# Packages: janitor

cholera_linelist %>% 
  tabyl(sex, outcome) # two arguments separated by a comma is used to build a 2-way table
```

## Advanced multi-adorned table

By chaining adorn\_ functions, we can format a raw table into a highly legible, professional descriptive summary featuring nested row percentages alongside raw frequencies.

```{r}
# Packages: janitor

cholera_linelist %>% 
tabyl(sex, outcome) %>%                 # 1) build a two-way table: rows = sex, columns = outcome
  adorn_totals("row") %>%               # 2) add a “Total” row summing across outcomes
  adorn_percentages("row") %>%          # 3) turn counts into row-wise percentages
  adorn_pct_formatting(digits = 1) %>%  # 4) format percentages with one decimal & a “%” sign
  adorn_ns(position = "front")          # 5) show the raw counts (n) before each percentage
```

+------------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                              |
+================================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                           |
|                                                                                                                                                |
| Please open **“Session 2: Summaries, Measures, Tables, & Epi-curves_Exercise”**. Once it opens:                                                |
|                                                                                                                                                |
| 1.  Load the packages listed at the top of the file.                                                                                           |
|                                                                                                                                                |
| 2.  Complete **Step 1: Load the Cleaned Data** and **Step 2: Descriptive Frequency Tables**                                                    |
|                                                                                                                                                |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **10 minutes** to finish. |
+------------------------------------------------------------------------------------------------------------------------------------------------+

# Summary statistics for numeric variables using the `get_summary_stats()` function from the `rstatix` package

Unlike categorical variables, truly continuous variables (e.g., age) are better summarized by measures of central tendency (mean, median) and dispersion (standard deviation, interquartile range). The `get_summary_stats()` function returns these metrics and fits naturally into a dplyr pipe `(%>%)`, making it easy to integrate with the rest of the workflow.

## Overall summary statistics for age

```{r}
# Packages: rstatix

cholera_linelist %>%
get_summary_stats(age, type = "common")  # compute the standard “common” set of descriptive stats (n, mean, SD, median, IQR, min, max) for the age variable
```

## Summary statistics using a single grouping variable

Sometimes we want to see whether a numeric variable (*age*) differs across categories (*outcome*). By grouping the data by *outcome* and then calling the `get_summary_stats()` function from the `rstatix` package, we obtain counts, means, medians, and other descriptive measures for each outcome level (recovered, hospitalized, or death), making it easy to compare age distributions across the groups.

```{r}
# Packages: rstatix

cholera_linelist %>% 
  group_by(outcome) %>% # split the data into separate groups by patient outcome
  get_summary_stats(age, type = "common")# and compute the standard “common” set of descriptive stats for each outcome group
```

## Summary statistics using multiple grouping variable

By adding nested grouping fields, you can review finer strata within the data—such as monitoring whether age trends vary when interacting with both biological sex and clinical outcome.

When a single grouping factor isn’t enough, you can group by two (or more) categorical variables to look at more granular patterns.

In this example we ask whether age differs across sex and clinical outcome. Chaining `group_by(sex, outcome)` with `get_summary_stats()` returns n, mean, median, SD, IQR, min, and max for each sex & outcome stratum:

- Female – Recovered

- Female – Hospitalized

- Male – Recovered

and so on—making it easy to spot age differences that might otherwise be hidden.

```{r}
# Packages: rstatix

cholera_linelist %>% 
  group_by(sex, outcome) %>%  # split data by every Sex × Outcome combination
  get_summary_stats(age, type = "common")# and compute the standard “common” set of descriptive stats for each outcome group
```

# Simple statistical tests

To determine if variations between patient groups represent random noise or statistically significant signals, we employ classical statistical hypothesis testing.

## Parametric Comparisons:

In statistics, a parametric method assumes that your data follow a known, mathematical distribution. Most commonly a normal (bell-shaped) distribution, summarized by its mean and standard deviation.

When these assumptions are (at least approximately) true, parametric tests squeeze more information out of the data and are usually more powerful than their non-parametric counterparts.

Here are a few ways you can check to see if your data are suitable for parametric tests:

1.  Look at a histogram or density plot
    - Does it roughly resemble a bell curve?
    - Mild skewness is fine, but pronounced skew or heavy tails signal trouble.
2.  Check summary statistics
    - Mean ≈ median and skewness near zero usually indicate symmetry. Large gaps hint at skew.
    - Very small sample sizes (n \< 30 per group) magnify the impact of non-normality.
3.  Calucalte Kurtosis
    - \|excess kurtosis\| \< 1 → close enough to normal for most purposes.
    - 1 – 2 → worth a closer look (check plots, maybe transform).
    - \> 2 or \< –2 → heavy-tailed or very flat; parametric tests may be unreliable unless sample sizes are large.
4.  Run a quick normality test
    - Shapiro–Wilk test
5.  Draw a Q–Q (quantile–quantile) plot
    - If the points lie close to the diagonal line, the data fit a normal distribution reasonably well.
    - Systematic curves away from the line suggest non-normality.

### Student's t-test

A t-test compares the means of a continuous variable between exactly two independent groups. This test assumes that the underlying continuous metric follows a roughly normal distribution.

Here, we evaluate whether the mean *age* differs significantly between the two independent groups within *sex*: male and female.

```{r}
# No packages used, just bse R 

t.test(age ~ sex, data = cholera_linelist)
```

**Formatted reporting table using the `gtsummary` package**

Instead of copy-pasting console output, the `tbl_summary()` function within the `gtsummary` package converts your data frame directly into a publication-ready table. When you chain `add_p()`, the function runs the appropriate statistical test and inserts the corresponding p-value, giving you a complete, polished summary in one pipeline.

```{r}
# Packages: gtsummary, tidyverse

cholera_linelist %>% 
  select(age, sex) %>%                 # keep only the variables needed for the summary
  tbl_summary(                         # create a {gtsummary} table
    statistic = age ~ "{mean} ({sd})", # show mean and SD for age in each group
    by = sex                           # split the summary by sex (Female vs. Male)
  ) %>% 
  add_p(age ~ "t.test")                # append a p-value from a two-sample t-test comparing ages between the two sex groups
```

### Analysis of Variance (ANOVA)

A one-way ANOVA generalizes the independent-samples t-test by testing whether the mean of a continuous variable differs across two or more independent groups—for example, comparing mean *age* among patient *outcome*: who recovered, were hospitalized, or died.

```{r}
# Packages: rstatix, tidyverse

cholera_linelist %>% anova_test(age ~ outcome)
```

**Formatted ANOVA summary table via the gtsummary package**

Instead of copy-pasting console output, the `tbl_summary()` function within the `gtsummary` package converts your data frame directly into a publication-ready table. When you chain `add_p()`, the function runs the appropriate statistical test and inserts the corresponding p-value, giving you a complete, polished summary in one pipeline.

```{r}
# Packages: gtsummary, tidyverse

cholera_linelist %>%
  select(age, outcome) %>%
  tbl_summary(
    by = outcome,
    statistic = list(age ~ "{mean} ({sd})") # Wrapped in list() for safety
  ) %>%
  add_p(test = all_continuous() ~ "oneway.test") # Uses base R's robust one-way ANOVA
```

## Non-Parametric Comparisons:

### Chi-squared test

Chi-squared tests are used to compare the distribution of a categorical variable between two or more groups.

**Non-Parametric Comparisons: Deviations from Normality**

When a continuous variable is highly skewed, highly irregular, or contains extreme outliers, parametric assumptions fail. In these situations, we switch to non-parametric rank-based tests which compare medians and distributions rather than means.

### **Two-Group Non-Parametric Comparison: Wilcoxon Rank-Sum / Mann-Whitney U**

The Wilcoxon rank-sum test assesses whether the distributions of a continuous variable differ significantly between two categorical groups without assuming a bell-curve distribution.

Here we will calculate a wilcoxon rank sum test for age and sex

```{r}
# no packages used, just base R

wilcox.test(age ~ sex, data = cholera_linelist)
```

**Formatted Wilcoxon summary table using the `gtsummary` package**

Instead of copy-pasting console output, the `tbl_summary()` function within the `gtsummary` package converts your data frame directly into a publication-ready table. When you chain `add_p()`, the function runs the appropriate statistical test and inserts the corresponding p-value, giving you a complete, polished summary in one pipeline.

Notice that for non-parametric data, we alter our reporting metrics **from** mean (sd) **to** median (IQR) to accurately represent the data's central tendency.

```{r}
# Packages: gtsummary, tidyverse

cholera_linelist %>% 
  select(age, sex) %>%                          # keep variables of interest
  tbl_summary(                                  # gtsummary: build a summary table
    statistic = age ~ "{median} ({p25}, {p75})",# specify what statistic to show (this is default so could remove)
    by = sex) %>%                               # specify the grouping variable
  add_p(age ~ "wilcox.test")                    # gtsummary: add a column of p-values using a Wilcoxon rank-sum (Mann–Whitney) test
```

### **Multi-Group Non-Parametric Comparison: Kruskal-Wallis Test**

The Kruskal-Wallis test is a non-parametric test used to compare the distributions of a continuous variable between a categorical variable with more than two groups when the data is not normally distributed.

In this case, we will compare the age distribution of cholera cases between different outcomes (e.g., recovered, died, hospitalised).

```{r}
# no packages used, just base R

kruskal.test(age ~ outcome, data = cholera_linelist)
```

**Formatted Kruskal-Wallis summary table using the `gtsummary` package**

Instead of copy-pasting console output, the `tbl_summary()` function within the `gtsummary` package converts your data frame directly into a publication-ready table. When you chain `add_p()`, the function runs the appropriate statistical test and inserts the corresponding p-value, giving you a complete, polished summary in one pipeline.

Notice that for non-parametric data, we alter our reporting metrics **from** mean (sd) **to** median (IQR) to accurately represent the data's central tendency.

```{r}
# Packages: gtsummary, tidyverse

cholera_linelist %>% 
  select(age, outcome) %>%                       # keep variables of interest
  tbl_summary(                                   # gtsummary: build a summary table
    statistic = age ~ "{median} ({p25}, {p75})", # specify what statistic to show (this is default so could remove)
    by = outcome) %>%                            # specify the grouping variable
  add_p(age ~ "kruskal.test")      # gtsummary: add a column of p-values using a Kruskal-Wallis (for ≥2-group) non-parametric test
```

### **Categorical Association: Pearson χ² Test of Independence**

The Pearson χ² test checks whether two categorical variables are independent—that is, whether the distribution of one variable changes across the categories of the other.

In our case, we use it to determine whether the proportion of males and females differs across the three outcome groups (recovered, hospitalised, died).

```{r}
# no packages used, just base R

chisq.test(
  x = cholera_linelist$sex,
  y = cholera_linelist$outcome
)
```

**Formatted χ² summary table using the `gtsummary` package**

Instead of copy-pasting console output, the `tbl_summary()` function within the `gtsummary` package converts your data frame directly into a publication-ready table. When you chain `add_p()`, the function runs the appropriate statistical test and inserts the corresponding p-value, giving you a complete, polished summary in one pipeline.

Notice that for non-parametric data, we alter our reporting metrics **from** mean (sd) **to** median (IQR) to accurately represent the data's central tendency.

```{r}
# Packages: gtsummary, tidyverse

cholera_linelist %>%                
  tbl_cross(                         # gtsummary: create a two-way table
    row   = sex,                     # rows = sex (e.g., Female, Male)
    col   = outcome,                 # columns = clinical outcome (Recovered, Hospitalised, Died)
    percent = "row"                  # display row-wise percentages (sums to 100% within each sex)
  ) %>% 
  add_p(test = "chisq.test") %>%     # attach a Pearson χ² p-value for the sex × outcome association wiht calcualted p-value
  bold_labels()                      # gtsummary: bold the row/column headers for readability
```

+------------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                              |
+================================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                           |
|                                                                                                                                                |
| Please open **“Session 2: Summaries, Measures, Tables, & Epi-curves_Exercise”**. Once it opens:                                                |
|                                                                                                                                                |
| - Complete **Step 3: Summary Statistics & Simple Hypothesis Testing**                                                                          |
|                                                                                                                                                |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **10 minutes** to finish. |
+------------------------------------------------------------------------------------------------------------------------------------------------+

## Table 1: Comprehensive Baseline Characteristics Table

In epidemiological publications and outbreak reports, "Table 1" is the standard layout used to summarize all recorded demographic, spatial, and clinical variables within a single comprehensive table, stratified by a key variable like clinical outcome.

The code block below automatically detects variable formats, assigns appropriate parametric or non-parametric tests based on data types, computes the statistical differences, and applies clean bold label layouts.

```{r}
# ── packages ─────────────────────────────────────────────
# dplyr, gtsummary, gt          
# gt comes with {gtsummary} for pretty output

# ── build Table 1: stratify by outcome (recovered vs death) ─────────────
tbl1 <- cholera_linelist %>% 
  # 1. Select the outcome and all your variables
  select(
    outcome, age, age_group, state_name, sex, dehydration, treatment, 
    water_source, watery_diarrhea, vomiting, leg_cramps, dehydration_signs, fever
  ) %>% 
  # 2. Convert the categorical variables to factors explicitly
  mutate(
    across(
      c(age_group, state_name, sex, dehydration, treatment, water_source, 
        watery_diarrhea, vomiting, leg_cramps, dehydration_signs, fever), 
      as.factor
    ),     
    outcome = factor(outcome, levels = c("recovered", "death", "hospitalised"))   
  ) %>% 
  # 3. Create the summary table (gtsummary will auto-detect 'age' as continuous)
  tbl_summary(
    by        = outcome,
    statistic = list(
                 all_continuous()  ~ "{mean} ± {sd}",     
                 all_categorical() ~ "{n} ({p}%)"
               ),
    missing   = "no"                                     
  ) %>% 
  add_n() %>%  
  add_p(
    test = list(
      all_categorical() ~ "chisq.test.no.correct",
      all_continuous()  ~ "aov"
    )
  ) %>% 
  bold_labels() %>%  
  modify_spanning_header(all_stat_cols() ~ "**Outcome**") %>%
  modify_header(label = "**Variable**")

# ── display in notebook / Jupyter / Quarto ───────────────
tbl1
```

+------------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                              |
+================================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                           |
|                                                                                                                                                |
| Please open **“Session 2: Summaries, Measures, Tables, & Epi-curves_Exercise”**. Once it opens:                                                |
|                                                                                                                                                |
| - Complete **Step 4: Constructing a Publication-Ready "Table 1"**                                                                              |
|                                                                                                                                                |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **15 minutes** to finish. |
+------------------------------------------------------------------------------------------------------------------------------------------------+

# Common epidemiological measures used for case-only data

## Case Fatality Risk / Ratio (CFR)

The case-fatality ratio (or risk) is the proportion of identified cases of a disease that result in death—a straightforward gauge of how severe the illness is and how well case management and surveillance are working.

**Formula**\
CFR (%) = (Deaths ÷ Total cases) × 100

**Interpretation & Use**

- Monitor it continuously during outbreaks to detect changes in disease severity or gaps in care.

- For example, cholera, WHO recommends a CFR below 1 %. Values rising to 3 – 4 % signal serious problems—such as delayed care-seeking, inadequate rehydration supplies, or other failures in the health-care system—that demand immediate action.

```{r}
# Packages: tidyverse, dplyr

cholera_linelist %>%
  summarise(
    total_cases = n(),
    total_deaths = sum(outcome == "death", na.rm = TRUE),
    cfr_percent = (total_deaths / total_cases) * 100
  )
```

## Proportional Morbidity (Exposure Proportions)

**Definition**\
When the size of the population at risk is unknown, epidemiologists often describe the distribution of specific exposures among the people who have already become ill. The proportion of cases reporting a given exposure indicates how common that exposure is within the case group.

**Formula**\
Proportion of cases with exposure X (%) = (Cases with exposure X ÷ Total cases) × 100

**Purpose**\
Tracking these proportions over time helps identify shifts in likely sources of infection. For example, if 80 % of cases in week 1 report drinking piped water but, by week 3, 75 % report eating at a particular market stall, investigators can focus control measures on the emerging exposure.

```{r}
# Packages: janitor, tidyverse
# Tracks the changing proportion of water sources among cases over time

cholera_linelist %>%
  #filter( !is.na(onset_date) & onset_date >= as.Date("2025-01-01") & onset_date <= as.Date("2025-01-31")) %>% #can resitrct time
  tabyl(water_source) %>%              # count cases by reported water source
  adorn_pct_formatting()               # add a column showing counts as %
```

+------------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                              |
+================================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                           |
|                                                                                                                                                |
| Please open **“Session 2: Summaries, Measures, Tables, & Epi-curves_Exercise”**. Once it opens:                                                |
|                                                                                                                                                |
| - Complete **Step 5: Case Fatality Risk / Ratio (CFR)** and **Step 6: Proportional Morbidity (Exposure Proportions)**                          |
|                                                                                                                                                |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **10 minutes** to finish. |
+------------------------------------------------------------------------------------------------------------------------------------------------+