---
title: "Session 1 - Introduction to Programming in R for Epidemiology." 
subtitle: "Part 3: Tidyverse, importing data, data cleaning and validation"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

In this session we will cover data importing as well as cleaning and validation with Tidyverse using a cholera line-list dataset for Sudan. Once we have cleaned our data, we will export it to the "processed_data" folder.

## Exercises

Hands-on practice breaks are built into the lesson. The exercise file for this session is located in `student_code/` and is named:\
“Session 1 – Part 3 – Tidyverse_importing-data_data-cleaning-and-validation_Exercise.qmd”

## Answer Key

Solutions are in the same folder. The answer sheet for this session is:\
“ANSWERS_Session 1 – Tidyverse, importing-data, data-cleaning-and-validation_Exercise.qmd”

# What are R Packages

- Packages are collections of functions and data that extend the functionality of the base R programming language.

- They are used to perform specialized tasks, from creating graphs to running complex statistical models.

- Packages are a core reason for R's popularity, as they are created and shared by a global community of users.

## Installing packages

You can easily install (most of) packages using the install.packages() function directly in R. For example:

```{r}
install.packages("tidyverse")
```

## Loading packages

After installing a package, you need to load it into your R session using the library() function. For example:

```{r}
library(tidyverse)
```

## Package management with pacman

The pacman package provides a convenient way to install and load multiple packages at once. It checks if the package is already installed, installs it if necessary, and then loads it. We will see how we use it for this exercise now int he next session.

# Packages used for this session

```{r}
pacman::p_load(
  # ── Core Infrastructure & Data I/O ─────────────────────────────────────────
  here,         # Relative file pathways to prevent broken working directory links
  rio,          # Swiss-Army knife for importing/exporting data (xlsx, csv, rds)
  tidyverse,    # Core ecosystem (ggplot2, dplyr, tidyr) for data science & viz

  # ── Data Cleaning & Standardisation ────────────────────────────────────────
  janitor,      # Clean column names, find duplicates, and build quick tabyls
  lubridate,    # Parsing and executing calculations on epidemic date vectors
  matchmaker,   # Dictionary-based string cleaning and surveillance recoding

# ── Field Epidemiology & Summaries ─────────────────────────────────────────
  epikit,       # Inline tracking functions and quick age_categories() generation
  skimr         # Fast, summary summaries of missingness and data distributions
)
```

# Importing data into R using Rio() and Here()

## Quick background

### Base R (no packages)

- This is the original, old school way of reading `.csv`, `.txt`, and `.dat` files and returns a data frame in the R environment using the function read.*csv,* read.*table*, etc.

- Cumbersome - it requires additional manual coding specifications based on the file type; it can be slow for large datasets; and can lead to unexpected behaviors with certain column types (i.e. converting character columns into factors)

### readr package

- Modern alternative to *Base R* and requires less configuration while being faster, cleaner, and provides less coding headaches (its does not convert character columns into factors) and uses the function read\_*csv,* read\_*tsv,* etc.

- Returns a *tibble,* a modern data frame with features that make it print nicely, displays on a screen better, does not change variables names, and does not use row labels. The output when inspecting the data set using *str()* are nicer looking and more informative

### haven package

- The **`haven`** package is the dedicated tool for reading proprietary formats like SPSS (`.sav`), SAS (`.xpt` & `.sas7bdat`), and Stata (`.dta`), and is a key part of the Tidyverse ecosystem.

### openxlsx package

- The **`openxlsx`** package is the dedicated tool for reading and writing data from **Excel files** (`.xlsx`) without needing external dependencies like Java. It is widely used for working with multi-sheet workbooks, applying styles, and handling complex Excel features, offering functions like `read.xlsx()` for import. 

## Why use the `rio` and `here` packages instead of base R?

### rio package

- Is highly useful because it provides a single, **universal function** for importing virtually any data file format.

- The `import()` function automatically **detects the file type** based on the file extension (e.g., `.csv`, `.xlsx`, `.RData`, `.rds`, `.xpt`, `.sas7bdat`, `.dta`, `.sav`, `.json,.txt`, `.dat`) and uses the appropriate specialized package (`readr`,`haven`, and `openxlsx`) to import it.

### here package

- The package **here** and its function `here()` make it easy to tell R where to find and to save your files - in essence, it builds file paths.

- Using **here()** works with you **R Project** to eliminate path problems -\> Relative paths = good for sharing; Absolute paths = Bad for sharing

  - **Bad:** `"/Users/mwegener/Desktop/maybe_here/data.csv"`

  - **Good:** `here("raw_data", "data.csv")`

- This is useful when the R project may be shared or accessed by multiple people/computers

```{r}
# if your .xlsx has multiple sheets:
# my_data <- import(here("FOLDER NAME", "DATA_FILE.xlsx"), which = "SHEET_NAME")
# packages used: rio, here, openxlsx

cholera_linelist_raw <- import(here('raw_data', 'cholera_linelist_2025_sudan.xlsx'))
```

+-----------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                             |
+===============================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                          |
|                                                                                                                                               |
| Please open **“Session 1: Tidyverse, importing data, data cleaning and validation_Exercise”**. Once it opens:                                 |
|                                                                                                                                               |
| 1.  Load the packages listed at the top of the file.                                                                                          |
|                                                                                                                                               |
| 2.  Complete **Step 1: Importing Data**.                                                                                                      |
|                                                                                                                                               |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **5 minutes** to finish. |
+-----------------------------------------------------------------------------------------------------------------------------------------------+

# Data cleaning and validation

<https://epirhandbook.com/en/new_pages/cleaning.html>

## The Tidyverse

### **A Unified Data Science Ecosystem**

The **tidyverse** is a collection of R packages designed around shared principles for efficient, readable data analysis. Install all packages at once with `install.packages("tidyverse")` and load with `library(tidyverse)`.

### **Core Philosophy:**

- **Tidy data structure**: Each variable = column, each observation = row, each value = cell

- **Consistent syntax**: Predictable function behavior across all packages

- **Human-centered design**: Intuitive functions that reduce learning curve

- **Pipe workflows**: Use `%>%` (`|>`) to chain operations for readable, step-by-step analysis

### **Key Benefits:**

- **Uniform data structures** using data frames/tibbles

- **Reproducible analysis** with version control emphasis

- **Clean naming conventions** (lowercase_with_underscores)

- **Active community** support and best practices

- **Functional pipelines** that transform data through logical steps

### Core Functions

+-------------------------------------------------------+--------------------------------------------------+-----------------+
| Function                                              | Utility                                          | Package         |
+=======================================================+==================================================+=================+
| `%>%`                                                 | “Pipe” (pass data from one function to the next) | magrittr        |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `mutate()`                                            | Create, transform, and redefine columns          | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `select()`                                            | Keep, remove, select, or rename columns          | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `rename()`                                            | Rename columns                                   | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `clean_names()`                                       | Standardize column-name syntax                   | janitor         |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `as.character()`, `as.numeric()`, `as.Date()`, *etc.* | Convert the class of a column                    | base R          |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `across()`                                            | Transform multiple columns at once               | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| *tidyselect* helpers                                  | Use logic to select columns                      | tidyselect      |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `filter()`                                            | Keep certain rows                                | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `distinct()`                                          | De-duplicate rows                                | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `rowwise()`                                           | Perform operations within each row               | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `add_row()`                                           | Add rows manually                                | tibble          |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `arrange()`                                           | Sort rows                                        | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `recode()`                                            | Recode values in a column                        | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `case_when()`                                         | Recode values with complex logical criteria      | dplyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `replace_na()`, `na_if()`, `coalesce()`               | Handle missing values                            | tidyr           |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `age_categories()`, `cut()`                           | Create categorical groups from a numeric column  | epikit / base R |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `match_df()`                                          | Recode/clean values using a data dictionary      | matchmaker      |
+-------------------------------------------------------+--------------------------------------------------+-----------------+
| `which()`                                             | Apply logical criteria; return indices           | base R          |
+-------------------------------------------------------+--------------------------------------------------+-----------------+

## Step 1: inspect the data

### Inspecting data helps identify potential problems before analysis and plan your data cleaning strategy by verifying:

- Data imported correctly

- Variable types are appropriate

- Value formatting is consistent

- Missing value patterns

- Duplicate records

- Data set structure and dimensions

- **If column names are clean (concise, no spaces, no blanks, no dots, no caps)**

```{r}
# Using head 
head(cholera_linelist_raw)
```

```{r}
#from skimr package
skim(cholera_linelist_raw)
```

```{r}
# from glimpse package
glimpse(cholera_linelist_raw)
```

**Now let's check character strings to see if:**

- Duplicate entries for the same value due to spelling errors

- inconsistent capitalization

- extra spaces

- if there are more values what you were expecting

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(State)
```

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(source water)
```

### what issues did you identify?

+----------------+-----------------------------------------------------------------------------------------------------------------+
| Issue          | Detail / Example                                                                                                |
+================+=================================================================================================================+
| Column names   | Mixed conventions—capitals, spaces, dots.                                                                       |
+----------------+-----------------------------------------------------------------------------------------------------------------+
| Date columns   | Stored as POSIXct; convert to Date for consistency.                                                             |
+----------------+-----------------------------------------------------------------------------------------------------------------+
| State          | Some entries are misspelled.                                                                                    |
+----------------+-----------------------------------------------------------------------------------------------------------------+
| age            | Column is character, should be numeric.                                                                         |
+----------------+-----------------------------------------------------------------------------------------------------------------+
| Symptom fields | Each symptom column contains missing values.                                                                    |
+----------------+-----------------------------------------------------------------------------------------------------------------+
| Code error     | `cholera_linelist_raw %>% count(source water)` fails because the column name contains a space (`source water`). |
+----------------+-----------------------------------------------------------------------------------------------------------------+

## Step 2: Clean data using Tiduverse

Using **Tidyverse** packages, we build one single cleaning pipeline using the pipe operator (`%>%`). Each pipe passes the result of one step to the next.

### 💡 **Pro Tip:**

We assign the result to a **new** data frame (`cholera_linelist_clean`) using `<-` to preserve the original raw data — always protect your raw data!

### We are going to do the following cleaning steps in one pipeline:

#### 2a. Standardize Column Names

`clean_names()` from **janitor** automatically converts all column names to `snake_case` (e.g., `date onset` → `date_onset`).

### 2b — Convert Date Columns to Date Format

`mutate()` + `as.Date()` converts "date onset" from POSIXct to an easier to use general date variable — required for any sorting, filtering, or date math.

### 2c — Rename Columns

`rename()` gives three columns clearer, more descriptive names:

| Old Name       | New Name       |
|----------------|----------------|
| `date_onset`   | `onset_date`   |
| `lat`          | `latitiude`    |
| `lon`          | `longitude`    |
| `source_water` | `water_source` |

### 2d — Standardise categorical text (“State” and “Sex”)

Many surveillance variables are free-text. Typos, mixed capitals, or different languages make analysis impossible until they are harmonized. We tackle two of them:

1.  **State** – turn “Blue Nile”, “Blu Nile”, “blue_nile”… into a single value “Blue Nile”.
2.  **Sex** – collapse “M”, “Male ”, “male”, “F ”, “Female”, blanks, etc. into the tidy set of values (male, female, unknown).

We use mutate() + case_when() (with stringr::str_detect() for pattern matching). The first matching rule wins; anything that matches nothing (including NA or empty strings) falls through to "unknown".

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(State)
```

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(SEX)
```

### 2e — Create Epidemiologically Meaningful Age Groups

Age is numeric in the raw line list, but most descriptive analyses and visualizations of cholera data require age strata (e.g., to calculate age-specific attack rates or draw epidemic curves by age band).

\
In this step we transform the single “age” column into an ordered factor called **age_group** with six mutually exclusive categories:

• 0–4 • 5–14 • 15–29 • 30–44 • 45–59 • 60 +

Why group ages?

• Interpretability – health audiences more easily grasp “cases among children 0–4 yrs” than a mean age of 6.3.\
• Comparability – standard WHO/UNICEF age bands match other outbreak reports and surveillance systems.\
• Stability – categorical rates are less sensitive to heaping or single-year mis-reporting.\
• Privacy – grouping reduces the risk of identifying very old or very young individuals in small datasets.

Implementation with the tidyverse:

1.  `mutate()` adds a new variable without altering the raw `age`. 

2.  `cut()` bins ages using explicit break points: `c(0, 5, 15, 30, 45, 60, Inf)`. 

3.  `right = FALSE` produces half-open intervals `[a, b)` so age 5 falls in “5–14”, not “0–4”. 

4.  `include.lowest = TRUE` ensures newborns (age 0) are included in the first band. 

5.  Custom `labels = c("0–4", …, "60+")` give reader-friendly category names. 

6.  Because `cut()` returns a factor ordered by the break sequence, plots and tables will display age groups in the correct epidemiological order without extra sorting code.

### 2f — Handling missing values in the symptom variables

For surveillance data we rarely delete records; instead we make missingness explicit so it is counted and reported.\
In this project we recode every blank or `NA` symptom entry to the literal value **"Unknown"**.

Why this approach?\
• “Unknown” appears in tables and plots, reminding users that information was not recorded.\
• Numeric columns stay unchanged, so analytical calculations (rates, medians, etc.) are unaffected.\
• The pipeline is fully reproducible and works for any future dataset with the same structure.

Steps implemented with the tidyverse:

1.  `mutate()` updates the data frame in place. 

2.  `across()` applies the same rule to all text variables. 

3.  `where(is.character)` limits the change to character columns only. 

4.  `trimws()` strips invisible spaces, `na_if()` turns empty strings into `NA`, and `replace_na()` converts those `NA` (plus any that were already present) to `"Unknown"`.

### 2g — Select and Order Final Variables

`select()` keeps only the necessary variables and arranges them in a logical order for interpretation before exporting the finalized dataset. You can also drop specific variables using the `-` sign, for example `select(-state)` would drop the original state variable since we have created a new one with the correct names.

### Start cleaning

This code chunk contains the complete workflow for cleaning the dataset, and you can run all the code between the pipes `%>%`.

```{r}
# packages used: janitor, dplyr, stringr, tidyr, tidyverse

cholera_linelist_clean <- cholera_linelist_raw %>%
  # STEP 2a: Standardize all column names to snake_case (lowercase with underscores)
  # Example: "Date Of Onset" becomes "date_of_onset"
  clean_names() %>%

  # STEP 2b: Convert date columns from character strings to proper R Date objects
  # This is required before doing any date math or sorting by date
  mutate(date_onset = as.Date(date_onset)) %>%
  # STEP 2c: Rename columns to be more descriptive and consistent
  # new name (left)  →  Old name (right)
  # onset_date   →  date_onset
  # latitude     →  lat
  # longitude   →  lon
  # water_source → source_water
  rename(
    onset_date   = date_onset,
    latitude     = lat,
    longitude   = lon,
    water_source = source_water 
  ) %>%
  # STEP 2d: Standardise messy categorical text
  # State  → new variable `state_name`
  #     - str_detect() searches the raw `state` values for spelling fragments
  #     - case_when() applies the first matching rule, else “Other/Unknown”
  # Sex    → overwrites `sex`
  #     - handles “M/m/Male/male…”, “F/f/Female/female…”, blanks, and NA
  #     - maps everything to lowercase `male`, `female`, or `unknown`
  #   str_detect() does the pattern matching;
  #   case_when() chains the if/else logic (first match wins).
 mutate(
    # ----- State -------------------------------------------------------
    state_name = case_when(
      str_detect(state, regex("blu|blue",  ignore_case = TRUE)) ~ "Blue Nile",
      str_detect(state, regex("gez",       ignore_case = TRUE)) ~ "Gezira",
      str_detect(state, regex("kass",      ignore_case = TRUE)) ~ "Kassala",
      str_detect(state, regex("^khar",     ignore_case = TRUE)) ~ "Khartoum",
      str_detect(state, regex("north",     ignore_case = TRUE)) ~ "North Darfur",
      str_detect(state, regex("red",       ignore_case = TRUE)) ~ "Red Sea",
      str_detect(state, regex("south",     ignore_case = TRUE)) ~ "South Darfur",
      str_detect(state, regex("white",     ignore_case = TRUE)) ~ "White Nile",
      is.na(state) | str_trim(state) == ""                       ~ "Other/Unknown",
      TRUE                                                       ~ "Other/Unknown"
    ),
    
    # ----- Sex ---------------------------------------------------------
    sex = case_when(
      str_detect(sex, regex("^m$",      ignore_case = TRUE)) ~ "male",
      str_detect(sex, regex("^male$",   ignore_case = TRUE)) ~ "male",
      str_detect(sex, regex("^f$",      ignore_case = TRUE)) ~ "female",
      str_detect(sex, regex("^female$", ignore_case = TRUE)) ~ "female",
      is.na(sex) | str_trim(sex) == ""                       ~ "unknown",
      TRUE                                                   ~ "unknown"
    )
  ) %>% 
  
  # STEP 2e ── Derive ordered age-group factor ---------------------------
  mutate(
    age_group = cut(
      age,
      breaks = c(0, 5, 15, 30, 45, 60, Inf),   # interval end-points
      right   = FALSE,                         # left-closed, right-open [a, b)
      include.lowest = TRUE,
      labels = c("0–4", "5–14", "15–29",
                 "30–44", "45–59", "60+")
    )
  ) %>%
  # STEP 2f: Change blanks or NA values to "unknown" 
  mutate(
    across(
      where(is.character),
      ~ replace_na( na_if(trimws(.x), ""), "unknown" )
    )
  ) %>%
  # STEP 2g: Select only the variables we need for analysis and arrange them in
  select(case_id, onset_date, state_name, age_group, everything(), -state)
```

+------------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                              |
+================================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                           |
|                                                                                                                                                |
| Please open **“Session 1: Tidyverse, importing data, data cleaning and validation_Exercise”**. Once it opens:                                  |
|                                                                                                                                                |
| - Complete **Step 2: Data Cleaning using the tidyverse**                                                                                       |
|                                                                                                                                                |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **15 minutes** to finish. |
+------------------------------------------------------------------------------------------------------------------------------------------------+

### Reinspect your cleaned data

```{r}
# Let's inspect the state_name varaible
cholera_linelist_clean %>% count(state_name)
```

```{r}
# Let's inspect the sex varaible
cholera_linelist_clean %>% count(sex)
```

```{r}
# Using the skimr package to inspect the cleaned data
skim(cholera_linelist_clean)
```

# Export data to the "processed_data" folder using Rio() and Here()

In the same spirit that we *imported* raw files earlier (`rio::import(here("raw_data", …))`), we now *export* our cleaned line-list with two helper packages:

- **`rio`** – wraps dozens of I/O functions behind a single verb, `export()`. The file type is inferred from the extension you give it (`.csv`, `.xlsx`, `.rds`, etc.).

- **`here`** – builds file paths that always start at the project root, so your code runs on any machine without fiddling with the working directory.

```{r}
# export the cleaned line-list to an Excel file in the "processed_data" folder
# packages used: rio, here, openxlsx
export(cholera_linelist_clean, here('processed_data', 'cholera_linelist_clean.xlsx'))
```

+-----------------------------------------------------------------------------------------------------------------------------------------------+
| Exercise Breakout                                                                                                                             |
+===============================================================================================================================================+
| We’ll pause for a short breakout exercise. You’ll be placed into breakout rooms now.                                                          |
|                                                                                                                                               |
| Please open **“Session 1: Tidyverse, importing data, data cleaning and validation_Exercise”**. Once it opens:                                 |
|                                                                                                                                               |
| - Complete **Step 3: Reinspect the cleaned dataframe** and **Step 4: Exporting Cleaned Data**                                                 |
|                                                                                                                                               |
| You can work on this together in your breakout room—talk through the steps and problem-solve as a group. You’ll have **5 minutes** to finish. |
+-----------------------------------------------------------------------------------------------------------------------------------------------+