---
title: "Session 1: Tidyverse, importing data, data cleaning and validation"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

qfqfqq

# Packages used for this session

```{r}
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  matchmaker, # dictionary-based cleaning
  epikit,     # age_categories() function
  tidyverse,  # data management and visualization
  skimr
)
```

# Importing data into R using Rio() and Here()

## Quick background

### Base R

-   This is the original, old school way of reading `.csv`, `.txt`, and `.dat` files and returns a data frame in the R environment using the function read.*csv,* read.*table*, etc.

-   Cumbersome - it reacquires additional manual coding specifications based on the file type; it can be slow for large data sets; and can lead to unexpected behaviors with certain column types (i.e. converting character columns into factors)

### readr()

-   Modern alternative to *Base R* and requires less configuration while being faster, cleaner, and provides less coding headaches (its does not convert character columns into factors) and uses the fucntion read\_*csv,* read\_*tsv,* etc.

-   Returns a *tibble,* a modern data frame with features that make it print nicely, displays on a screen better, does not change variables names, and does not use row labels. The output when inspecting the data set using *str()* are nicer looking and more informative

### haven()

-   The **`haven`** package is the dedicated tool for reading proprietary formats like SPSS (`.sav`), SAS (`.xpt` & `.sas7bdat`), and Stata (`.dta`), and is a key part of the Tidyverse ecosystem.

### openxlsx()

-   The **`openxlsx`** package is the dedicated tool for reading and writing data from **Excel files** (`.xlsx`) without needing external dependencies like Java. It is widely used for working with multi-sheet workbooks, applying styles, and handling complex Excel features, offering functions like `read.xlsx()` for import. 

## Why use rio() andf here()n instead of base R?

### rio()

-   Is highly useful because it provides a single, **universal function** for importing virtually any data file format.

-   The `import()` function automatically **detects the file type** based on the file extension (e.g., `.csv`, `.xlsx`, `.RData`, `.rds`, `.xpt`, `.sas7bdat`, `.dta`, `.sav`, `.json,.txt`, `.dat`) and uses the appropriate specialized package (`readr`,`haven`, and `openxlsx`) to import it.

### here()

-   The package **here** and its function `here()` make it easy to tell R where to find and to save your files - in essence, it builds file paths.

-   Using **here()** works with you **R Project** to eliminate path roblems -\> Relative paths = good for sharing; Absolute paths = Bad for sharing

    -   **Bad:** `"/Users/mwegener/Desktop/maybe_here/data.csv"`

    -   **Good:** `here("raw_data", "data.csv")`

-   This is useful when the R project may be shared or accessed by multiple people/computers

```{r}
# if your .xlsx has multiple sheets:
# my_data <- import(here("FOLDER NAME", "DATA_FILE.xlsx"), which = "SHEET_NAME")

cholera_linelist_raw <- import(here('raw_data', 'cholera_linelist_2025_sudan.csv'))
```

# Data cleaning and validation

<https://epirhandbook.com/en/new_pages/cleaning.html>

## The Tidyverse

### **A Unified Data Science Ecosystem**

The **tidyverse** is a collection of R packages designed around shared principles for efficient, readable data analysis. Install all packages at once with `install.packages("tidyverse")` and load with `library(tidyverse)`.

### **Core Philosophy:**

-   **Tidy data structure**: Each variable = column, each observation = row, each value = cell

-   **Consistent syntax**: Predictable function behavior across all packages

-   **Human-centered design**: Intuitive functions that reduce learning curve

-   **Pipe workflows**: Use `%>%` to chain operations for readable, step-by-step analysis

### **Key Benefits:**

-   **Uniform data structures** using data frames/tibbles

-   **Reproducible analysis** with version control emphasis

-   **Clean naming conventions** (lowercase_with_underscores)

-   **Active community** support and best practices

-   **Functional pipelines** that transform data through logical steps

### Core Functions

| Function | Utility | Package |
|----|----|----|
| `%>%` | “pipe” (pass) data from one function to the next | **magrittr** |
| `mutate()` | create, transform, and re-define columns | **dplyr** |
| `select()` | keep, remove, select, or re-name columns | **dplyr** |
| `rename()` | rename columns | **dplyr** |
| `clean_names()` | standardize the syntax of column names | **janitor** |
| `as.character()`, `as.numeric()`, `as.Date()`, etc. | convert the class of a column | **base** R |
| `across()` | transform multiple columns at one time | **dplyr** |
| **tidyselect** functions | use logic to select columns | **tidyselect** |
| `filter()` | keep certain rows | **dplyr** |
| `distinct()` | de-duplicate rows | **dplyr** |
| `rowwise()` | operations by/within each row | **dplyr** |
| `add_row()` | add rows manually | **tibble** |
| `arrange()` | sort rows | **dplyr** |
| `recode()` | re-code values in a column | **dplyr** |
| `case_when()` | re-code values in a column using more complex logical criteria | **dplyr** |
| `replace_na()`, `na_if()`, `coalesce()` | special functions for re-coding | **tidyr** |
| `age_categories()`and `cut()` | create categorical groups from a numeric column | **epikit** and **base** R |
| `match_df()` | re-code/clean values using a data dictionary | **matchmaker** |
| `which()` | apply logical criteria; return indices | **base** R |

## Step 1: inspect the data

### Inspecting data helps identify potential problems before analysis and plan your data cleaning strategy by verifying:

-   Data imported correctly

-   Variable types are appropriate

-   Value formatting is consistent

-   Missing value patterns

-   Duplicate records

-   Data set structure and dimensions

-   **If column names are clean (concise, no spaces, no blanks, no dots, no caps)**

```{r}
#from skimr package
skim(cholera_linelist_raw)
```

### what issues did you identify?

## Step 2: 
