---
title: "Session 1: Tidyverse, importing data, data cleaning and validation"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

qfqfqq

# Packages used for this session

```{r}
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  matchmaker, # dictionary-based cleaning
  epikit,     # age_categories() function
  tidyverse,  # data management and visualization
  skimr
)
```

# Importing data into R using Rio() and Here()

## Quick background

### Base R

- This is the original, old school way of reading `.csv`, `.txt`, and `.dat` files and returns a data frame in the R environment using the function read.*csv,* read.*table*, etc.

- Cumbersome - it reacquires additional manual coding specifications based on the file type; it can be slow for large data sets; and can lead to unexpected behaviors with certain column types (i.e. converting character columns into factors)

### readr()

- Modern alternative to *Base R* and requires less configuration while being faster, cleaner, and provides less coding headaches (its does not convert character columns into factors) and uses the fucntion read\_*csv,* read\_*tsv,* etc.

- Returns a *tibble,* a modern data frame with features that make it print nicely, displays on a screen better, does not change variables names, and does not use row labels. The output when inspecting the data set using *str()* are nicer looking and more informative

### haven()

- The **`haven`** package is the dedicated tool for reading proprietary formats like SPSS (`.sav`), SAS (`.xpt` & `.sas7bdat`), and Stata (`.dta`), and is a key part of the Tidyverse ecosystem.

### openxlsx()

- The **`openxlsx`** package is the dedicated tool for reading and writing data from **Excel files** (`.xlsx`) without needing external dependencies like Java. It is widely used for working with multi-sheet workbooks, applying styles, and handling complex Excel features, offering functions like `read.xlsx()` for import. 

## Why use rio() andf here()n instead of base R?

### rio()

- Is highly useful because it provides a single, **universal function** for importing virtually any data file format.

- The `import()` function automatically **detects the file type** based on the file extension (e.g., `.csv`, `.xlsx`, `.RData`, `.rds`, `.xpt`, `.sas7bdat`, `.dta`, `.sav`, `.json,.txt`, `.dat`) and uses the appropriate specialized package (`readr`,`haven`, and `openxlsx`) to import it.

### here()

- The package **here** and its function `here()` make it easy to tell R where to find and to save your files - in essence, it builds file paths.

- Using **here()** works with you **R Project** to eliminate path roblems -\> Relative paths = good for sharing; Absolute paths = Bad for sharing

  - **Bad:** `"/Users/mwegener/Desktop/maybe_here/data.csv"`

  - **Good:** `here("raw_data", "data.csv")`

- This is useful when the R project may be shared or accessed by multiple people/computers

```{r}
# if your .xlsx has multiple sheets:
# my_data <- import(here("FOLDER NAME", "DATA_FILE.xlsx"), which = "SHEET_NAME")

cholera_linelist_raw <- import(here('raw_data', 'cholera_linelist_2025_sudan.xlsx'))
```

# Data cleaning and validation

<https://epirhandbook.com/en/new_pages/cleaning.html>

## The Tidyverse

### **A Unified Data Science Ecosystem**

The **tidyverse** is a collection of R packages designed around shared principles for efficient, readable data analysis. Install all packages at once with `install.packages("tidyverse")` and load with `library(tidyverse)`.

### **Core Philosophy:**

- **Tidy data structure**: Each variable = column, each observation = row, each value = cell

- **Consistent syntax**: Predictable function behavior across all packages

- **Human-centered design**: Intuitive functions that reduce learning curve

- **Pipe workflows**: Use `%>%` to chain operations for readable, step-by-step analysis

### **Key Benefits:**

- **Uniform data structures** using data frames/tibbles

- **Reproducible analysis** with version control emphasis

- **Clean naming conventions** (lowercase_with_underscores)

- **Active community** support and best practices

- **Functional pipelines** that transform data through logical steps

### Core Functions

+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| Function                                            | Utility                                                        | Package                   |
+=====================================================+================================================================+===========================+
| `%>%`                                               | “pipe” (pass) data from one function to the next               | **magrittr**              |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `mutate()`                                          | create, transform, and re-define columns                       | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `select()`                                          | keep, remove, select, or re-name columns                       | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `rename()`                                          | rename columns                                                 | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `clean_names()`                                     | standardize the syntax of column names                         | **janitor**               |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `as.character()`, `as.numeric()`, `as.Date()`, etc. | convert the class of a column                                  | **base** R                |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `across()`                                          | transform multiple columns at one time                         | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| **tidyselect** functions                            | use logic to select columns                                    | **tidyselect**            |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `filter()`                                          | keep certain rows                                              | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `distinct()`                                        | de-duplicate rows                                              | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `rowwise()`                                         | operations by/within each row                                  | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `add_row()`                                         | add rows manually                                              | **tibble**                |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `arrange()`                                         | sort rows                                                      | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `recode()`                                          | re-code values in a column                                     | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `case_when()`                                       | re-code values in a column using more complex logical criteria | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `replace_na()`, `na_if()`, `coalesce()`             | special functions for re-coding                                | **tidyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `age_categories()`and `cut()`                       | create categorical groups from a numeric column                | **epikit** and **base** R |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `match_df()`                                        | re-code/clean values using a data dictionary                   | **matchmaker**            |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `which()`                                           | apply logical criteria; return indices                         | **base** R                |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+

## Step 1: inspect the data

### Inspecting data helps identify potential problems before analysis and plan your data cleaning strategy by verifying:

- Data imported correctly

- Variable types are appropriate

- Value formatting is consistent

- Missing value patterns

- Duplicate records

- Data set structure and dimensions

- **If column names are clean (concise, no spaces, no blanks, no dots, no caps)**

```{r}
# Using head 
head(linelist_raw)
```

```{r}
#from skimr package
skim(cholera_linelist_raw)
```

```{r}
# from glimpse package
glimpse(cholera_linelist_raw)
```

**Now let's check character strings to see if:**

- Duplicate entries for the same value due to spelling errors

- inconsistent capitalization

- extra spaces

- if there are more values what you were expecting

```{r}
# Let's inspect the State varaible
cholera_linelist_raw count(State)
```

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(source water)
```

### what issues did you identify?

+--------------------------------------------------------------------------------------+
|                                                                                      |
+:=====================================================================================+
| Column Names: Mixed naming conventions, caps, spaces, dots                           |
+--------------------------------------------------------------------------------------+
| other dates are POSIXct but we can make them *Dates* for ease and consistency        |
+--------------------------------------------------------------------------------------+
| State: some are misspelled                                                           |
+--------------------------------------------------------------------------------------+
| age is shown as character value                                                      |
+--------------------------------------------------------------------------------------+
| missing data in every symptom                                                        |
+--------------------------------------------------------------------------------------+
| this code doe not work because there is a space in the variable name "source water"\ |
| \                                                                                    |
| cholera_linelist_raw %\>% count(source water)                                        |
+--------------------------------------------------------------------------------------+

## Step 2: Clean data using Tiduverse

Using **Tidyverse** packages, we build one single cleaning pipeline using the pipe operator (`%>%`). Each pipe passes the result of one step to the next.

### 💡 **Pro Tip:**

We assign the result to a **new** data frame (`cholera_linelist_clean`) using `<-` to preserve the original raw data — always protect your raw data!

### We are going to do the following cleaning steps in one pipeline:

#### 2a. Standardize Column Names

`clean_names()` from **janitor** automatically converts all column names to `snake_case` (e.g., `date onset` → `date_onset`).

### 2b — Convert Date Columns to Date Format

`mutate()` + `as.Date()` converts "date onset" from POSIXct to an easier to use general date variable — required for any sorting, filtering, or date math.

### 2c — Rename Columns

`rename()` gives three columns clearer, more descriptive names:

| Old Name       | New Name       |
|----------------|----------------|
| `date_onset`   | `onset_date`   |
| `lat`          | `latitiude`    |
| `lon`          | `longitutde`   |
| `source_water` | `water_soruce` |

### 2d — Standardize State Names

`case_when()` + `str_detect()` inside `mutate()` maps inconsistent raw State values to clean standardized names in a new `state_name` column. First match wins; anything unrecognized is labeled `"Other/Unknown"`.

We can use the code we ran previously as a reference to see what the correct spelling is:

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(State)
```

### 2e — Select and Order Final Variables

`select()` keeps only the necessary variables and arranges them in a logical order for interpretation before exporting the finalized dataset. You can also drop specific variables using the `-` sign, for emaple `select(-state)` would drop the original state variable since we have created a new one with the correct names.

### Start cleaning

This code chunk contains the complete workflow for cleaning the dataset, and you can run all the code between the pipes `%>%`.

```{r}

cholera_linelist_clean <- cholera_linelist_raw %>%
  # STEP 2a: Standardize all column names to snake_case (lowercase with underscores)
  # Example: "Date Of Onset" becomes "date_of_onset"
  clean_names() %>%

  # STEP 2b: Convert date columns from character strings to proper R Date objects
  # This is required before doing any date math or sorting by date
  mutate( date_onset = as.Date(date_onset)) %>%
  # STEP 2c: Rename columns to be more descriptive and consistent
  # new name (left)  →  Old name (right)
  # onset_date   →  date_onset
  # latitude     →  lat
  # longitutde   →  lon
  # water_source → source_water
  rename(
    onset_date   = date_onset,
    latitude     = lat,
    longitutde   = lon,
    water_source = source_water 
  ) %>%
   # STEP 2d: Standardize messy state name values into clean, consistent categories
   # str_detect() looks for a keyword pattern inside the raw hospital column
   # case_when() works like a series of if/else conditions — first match wins
   # A new varaible with the correct names is added and called "state_name"
    mutate(
      state_name = case_when(
        str_detect(state, "Blu|Blue") ~ "Blue Nile",
        str_detect(state, "Ge|Gez")   ~ "Gezira",
        str_detect(state, "Kas|Kass") ~ "Kassala",
        str_detect(state, "Khar")     ~ "Khartoum",
        str_detect(state, "North")    ~ "North Darfur",
        str_detect(state, "Red")      ~ "Red Sea",
        str_detect(state, "South")    ~ "South Darfur",
        str_detect(state, "White")    ~ "White Nile",
        is.na(state) | state == "" ~ "Other/Unknown",
        TRUE                             ~ "Other/Unknown"  # catches anything else
      )
    ) %>%
  # STEP 2e: Select only the variables we need for analysis and arrange them in
  select(case_id, onset_date, state_name, everything(), -state)
```

### Step 3: Group exercise

in groups of 4 or 5 do similar importing, inspecting, cleaning, and exporting for the dengue file in the raw_data fodler

Use the file in the student code folder for exercise 1...