---
title: "Session 1: Tidyverse, importing data, data cleaning and validation"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

In this session we will cover data improitng, cleaning, and validation using the Tidyverse using a cholera line-list dataset for Sudan. Once we have cleaned the data, we will export it to the "processed_data" folder. We will then do a group exercise where you will be doing the same thing with a dengue line-list dataset for Sudan.

# Packages used for this session

```{r}
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  matchmaker, # dictionary-based cleaning
  epikit,     # age_categories() function
  tidyverse,  # data management and visualization
  skimr
)
```

# Importing data into R using Rio() and Here()

## Quick background

### Base R

- This is the original, old school way of reading `.csv`, `.txt`, and `.dat` files and returns a data frame in the R environment using the function read.*csv,* read.*table*, etc.

- Cumbersome - it reacquires additional manual coding specifications based on the file type; it can be slow for large data sets; and can lead to unexpected behaviors with certain column types (i.e. converting character columns into factors)

### readr()

- Modern alternative to *Base R* and requires less configuration while being faster, cleaner, and provides less coding headaches (its does not convert character columns into factors) and uses the fucntion read\_*csv,* read\_*tsv,* etc.

- Returns a *tibble,* a modern data frame with features that make it print nicely, displays on a screen better, does not change variables names, and does not use row labels. The output when inspecting the data set using *str()* are nicer looking and more informative

### haven()

- The **`haven`** package is the dedicated tool for reading proprietary formats like SPSS (`.sav`), SAS (`.xpt` & `.sas7bdat`), and Stata (`.dta`), and is a key part of the Tidyverse ecosystem.

### openxlsx()

- The **`openxlsx`** package is the dedicated tool for reading and writing data from **Excel files** (`.xlsx`) without needing external dependencies like Java. It is widely used for working with multi-sheet workbooks, applying styles, and handling complex Excel features, offering functions like `read.xlsx()` for import. 

## Why use rio() andf here()n instead of base R?

### rio()

- Is highly useful because it provides a single, **universal function** for importing virtually any data file format.

- The `import()` function automatically **detects the file type** based on the file extension (e.g., `.csv`, `.xlsx`, `.RData`, `.rds`, `.xpt`, `.sas7bdat`, `.dta`, `.sav`, `.json,.txt`, `.dat`) and uses the appropriate specialized package (`readr`,`haven`, and `openxlsx`) to import it.

### here()

- The package **here** and its function `here()` make it easy to tell R where to find and to save your files - in essence, it builds file paths.

- Using **here()** works with you **R Project** to eliminate path roblems -\> Relative paths = good for sharing; Absolute paths = Bad for sharing

  - **Bad:** `"/Users/mwegener/Desktop/maybe_here/data.csv"`

  - **Good:** `here("raw_data", "data.csv")`

- This is useful when the R project may be shared or accessed by multiple people/computers

```{r}
# if your .xlsx has multiple sheets:
# my_data <- import(here("FOLDER NAME", "DATA_FILE.xlsx"), which = "SHEET_NAME")
# packages used: rio, here, openxlsx

cholera_linelist_raw <- import(here('raw_data', 'cholera_linelist_2025_sudan.xlsx'))
```

# Data cleaning and validation

<https://epirhandbook.com/en/new_pages/cleaning.html>

## The Tidyverse

### **A Unified Data Science Ecosystem**

The **tidyverse** is a collection of R packages designed around shared principles for efficient, readable data analysis. Install all packages at once with `install.packages("tidyverse")` and load with `library(tidyverse)`.

### **Core Philosophy:**

- **Tidy data structure**: Each variable = column, each observation = row, each value = cell

- **Consistent syntax**: Predictable function behavior across all packages

- **Human-centered design**: Intuitive functions that reduce learning curve

- **Pipe workflows**: Use `%>%` to chain operations for readable, step-by-step analysis

### **Key Benefits:**

- **Uniform data structures** using data frames/tibbles

- **Reproducible analysis** with version control emphasis

- **Clean naming conventions** (lowercase_with_underscores)

- **Active community** support and best practices

- **Functional pipelines** that transform data through logical steps

### Core Functions

+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| Function                                            | Utility                                                        | Package                   |
+=====================================================+================================================================+===========================+
| `%>%`                                               | “pipe” (pass) data from one function to the next               | **magrittr**              |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `mutate()`                                          | create, transform, and re-define columns                       | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `select()`                                          | keep, remove, select, or re-name columns                       | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `rename()`                                          | rename columns                                                 | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `clean_names()`                                     | standardize the syntax of column names                         | **janitor**               |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `as.character()`, `as.numeric()`, `as.Date()`, etc. | convert the class of a column                                  | **base** R                |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `across()`                                          | transform multiple columns at one time                         | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| **tidyselect** functions                            | use logic to select columns                                    | **tidyselect**            |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `filter()`                                          | keep certain rows                                              | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `distinct()`                                        | de-duplicate rows                                              | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `rowwise()`                                         | operations by/within each row                                  | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `add_row()`                                         | add rows manually                                              | **tibble**                |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `arrange()`                                         | sort rows                                                      | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `recode()`                                          | re-code values in a column                                     | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `case_when()`                                       | re-code values in a column using more complex logical criteria | **dplyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `replace_na()`, `na_if()`, `coalesce()`             | special functions for re-coding                                | **tidyr**                 |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `age_categories()`and `cut()`                       | create categorical groups from a numeric column                | **epikit** and **base** R |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `match_df()`                                        | re-code/clean values using a data dictionary                   | **matchmaker**            |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+
| `which()`                                           | apply logical criteria; return indices                         | **base** R                |
+-----------------------------------------------------+----------------------------------------------------------------+---------------------------+

## Step 1: inspect the data

### Inspecting data helps identify potential problems before analysis and plan your data cleaning strategy by verifying:

- Data imported correctly

- Variable types are appropriate

- Value formatting is consistent

- Missing value patterns

- Duplicate records

- Data set structure and dimensions

- **If column names are clean (concise, no spaces, no blanks, no dots, no caps)**

```{r}
# Using head 
head(cholera_linelist_raw)
```

```{r}
#from skimr package
skim(cholera_linelist_raw)
```

```{r}
# from glimpse package
glimpse(cholera_linelist_raw)
```

**Now let's check character strings to see if:**

- Duplicate entries for the same value due to spelling errors

- inconsistent capitalization

- extra spaces

- if there are more values what you were expecting

```{r}
# Let's inspect the State varaible
cholera_linelist_raw count(State)
```

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(source water)
```

### what issues did you identify?

+--------------------------------------------------------------------------------------+
|                                                                                      |
+:=====================================================================================+
| Column Names: Mixed naming conventions, caps, spaces, dots                           |
+--------------------------------------------------------------------------------------+
| other dates are POSIXct but we can make them *Dates* for ease and consistency        |
+--------------------------------------------------------------------------------------+
| State: some are misspelled                                                           |
+--------------------------------------------------------------------------------------+
| age is shown as character value                                                      |
+--------------------------------------------------------------------------------------+
| Every symptom has some missing data                                                  |
+--------------------------------------------------------------------------------------+
| this code doe not work because there is a space in the variable name "source water"\ |
| \                                                                                    |
| cholera_linelist_raw %\>% count(source water)                                        |
+--------------------------------------------------------------------------------------+

## Step 2: Clean data using Tiduverse

Using **Tidyverse** packages, we build one single cleaning pipeline using the pipe operator (`%>%`). Each pipe passes the result of one step to the next.

### 💡 **Pro Tip:**

We assign the result to a **new** data frame (`cholera_linelist_clean`) using `<-` to preserve the original raw data — always protect your raw data!

### We are going to do the following cleaning steps in one pipeline:

#### 2a. Standardize Column Names

`clean_names()` from **janitor** automatically converts all column names to `snake_case` (e.g., `date onset` → `date_onset`).

### 2b — Convert Date Columns to Date Format

`mutate()` + `as.Date()` converts "date onset" from POSIXct to an easier to use general date variable — required for any sorting, filtering, or date math.

### 2c — Rename Columns

`rename()` gives three columns clearer, more descriptive names:

| Old Name       | New Name        |
|----------------|-----------------|
| `date_onset`   | `onset_date`    |
| `lat`          | `latitiude`     |
| `lon`          | `longitutde`    |
| `source_water` | `water_sorucee` |

### 2d — Standardise categorical text (“State” and “Sex”)

Many surveillance variables are free-text. Typos, mixed capitals, or different languages make analysis impossible until they are harmonised. We tackle two of them:

1.  **State** – turn “Blue Nile”, “Blu Nile”, “blue_nile”… into a single value “Blue Nile”.
2.  **Sex** – collapse “M”, “Male ”, “male”, “F ”, “Female”, blanks, etc. into the tidy set of values (male, female, unknown).

We use mutate() + case_when() (with stringr::str_detect() for pattern matching). The first matching rule wins; anything that matches nothing (including NA or empty strings) falls through to "unknown".

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(State)
```

```{r}
# Let's inspect the State varaible
cholera_linelist_raw %>% count(SEX)
```

### 2e — Create Epidemiologically Meaningful Age Groups

Age is numeric in the raw line list, but most descriptive analyses and visualisations of cholera data require age strata (e.g., to calculate age-specific attack rates or draw epidemic curves by age band).\
In this step we transform the single “age” column into an ordered factor called **age_group** with six mutually exclusive categories:

• 0–4 • 5–14 • 15–29 • 30–44 • 45–59 • 60 +

Why group ages?

• Interpretability – health audiences more easily grasp “cases among children 0–4 yrs” than a mean age of 6.3.\
• Comparability – standard WHO/UNICEF age bands match other outbreak reports and surveillance systems.\
• Stability – categorical rates are less sensitive to heaping or single-year mis-reporting.\
• Privacy – grouping reduces the risk of identifying very old or very young individuals in small datasets.

Implementation with the tidyverse:

1.  `mutate()` adds a new variable without altering the raw `age`. 

2.  `cut()` bins ages using explicit break points: `c(0, 5, 15, 30, 45, 60, Inf)`. 

3.  `right = FALSE` produces half-open intervals `[a, b)` so age 5 falls in “5–14”, not “0–4”. 

4.  `include.lowest = TRUE` ensures newborns (age 0) are included in the first band. 

5.  Custom `labels = c("0–4", …, "60+")` give reader-friendly category names. 

6.  Because `cut()` returns a factor ordered by the break sequence, plots and tables will display age groups in the correct epidemiological order without extra sorting code.

### 2f — Handling missing values in the symptom variables

For surveillance data we rarely delete records; instead we make missingness explicit so it is counted and reported.\
In this project we recode every blank or `NA` symptom entry to the literal value **"Unknown"**.

Why this approach?\
• “Unknown” appears in tables and plots, reminding users that information was not recorded.\
• Numeric columns stay unchanged, so analytical calculations (rates, medians, etc.) are unaffected.\
• The pipeline is fully reproducible and works for any future dataset with the same structure.

Steps implemented with the tidyverse:

1.  `mutate()` updates the data frame in place. 

2.  `across()` applies the same rule to all text variables. 

3.  `where(is.character)` limits the change to character columns only. 

4.  `trimws()` strips invisible spaces, `na_if()` turns empty strings into `NA`, and `replace_na()` converts those `NA` (plus any that were already present) to `"Unknown"`.

### 2g — Select and Order Final Variables

`select()` keeps only the necessary variables and arranges them in a logical order for interpretation before exporting the finalized dataset. You can also drop specific variables using the `-` sign, for emaple `select(-state)` would drop the original state variable since we have created a new one with the correct names.

### Start cleaning

This code chunk contains the complete workflow for cleaning the dataset, and you can run all the code between the pipes `%>%`.

```{r}
# packages used: janitor, dplyr, stringr, tidyr, tidyverse

cholera_linelist_clean <- cholera_linelist_raw %>%
  # STEP 2a: Standardize all column names to snake_case (lowercase with underscores)
  # Example: "Date Of Onset" becomes "date_of_onset"
  clean_names() %>%

  # STEP 2b: Convert date columns from character strings to proper R Date objects
  # This is required before doing any date math or sorting by date
  mutate( date_onset = as.Date(date_onset)) %>%
  # STEP 2c: Rename columns to be more descriptive and consistent
  # new name (left)  →  Old name (right)
  # onset_date   →  date_onset
  # latitude     →  lat
  # longitutde   →  lon
  # water_source → source_water
  rename(
    onset_date   = date_onset,
    latitude     = lat,
    longitutde   = lon,
    water_source = source_water 
  ) %>%
  # STEP 2d: Standardise messy categorical text
  # State  → new variable `state_name`
  #     - str_detect() searches the raw `state` values for spelling fragments
  #     - case_when() applies the first matching rule, else “Other/Unknown”
  # Sex    → overwrites `sex`
  #     - handles “M/m/Male/male…”, “F/f/Female/female…”, blanks, and NA
  #     - maps everything to lowercase `male`, `female`, or `unknown`
  #   str_detect() does the pattern matching;
  #   case_when() chains the if/else logic (first match wins).
 mutate(
    # ----- State -------------------------------------------------------
    state_name = case_when(
      str_detect(state, regex("blu|blue",  ignore_case = TRUE)) ~ "Blue Nile",
      str_detect(state, regex("gez",       ignore_case = TRUE)) ~ "Gezira",
      str_detect(state, regex("kass",      ignore_case = TRUE)) ~ "Kassala",
      str_detect(state, regex("^khar",     ignore_case = TRUE)) ~ "Khartoum",
      str_detect(state, regex("north",     ignore_case = TRUE)) ~ "North Darfur",
      str_detect(state, regex("red",       ignore_case = TRUE)) ~ "Red Sea",
      str_detect(state, regex("south",     ignore_case = TRUE)) ~ "South Darfur",
      str_detect(state, regex("white",     ignore_case = TRUE)) ~ "White Nile",
      is.na(state) | str_trim(state) == ""                       ~ "Other/Unknown",
      TRUE                                                       ~ "Other/Unknown"
    ),
    
    # ----- Sex ---------------------------------------------------------
    sex = case_when(
      str_detect(sex, regex("^m$",      ignore_case = TRUE)) ~ "male",
      str_detect(sex, regex("^male$",   ignore_case = TRUE)) ~ "male",
      str_detect(sex, regex("^f$",      ignore_case = TRUE)) ~ "female",
      str_detect(sex, regex("^female$", ignore_case = TRUE)) ~ "female",
      is.na(sex) | str_trim(sex) == ""                       ~ "unknown",
      TRUE                                                   ~ "unknown"
    )
  ) %>% 
  
  # STEP 2e ── Derive ordered age-group factor ---------------------------
  mutate(
    age_group = cut(
      age,
      breaks = c(0, 5, 15, 30, 45, 60, Inf),   # interval end-points
      right   = FALSE,                         # left-closed, right-open [a, b)
      include.lowest = TRUE,
      labels = c("0–4", "5–14", "15–29",
                 "30–44", "45–59", "60+")
    )
  ) %>%
  # STEP 2f: Change blanks or NA values to "unknown" 
  mutate(
    across(
      where(is.character),
      ~ replace_na( na_if(trimws(.x), ""), "unknown" )
    )
  ) %>%
  # STEP 2g: Select only the variables we need for analysis and arrange them in
  select(   case_id, onset_date, state_name, age_group, everything(), -state)
```

### Reinspect your cleaned data

```{r}
# Let's inspect the state_name varaible
cholera_linelist_clean %>% count(state_name)
```

```{r}
# Let's inspect the sex varaible
cholera_linelist_clean %>% count(sex)
```

```{r}
# Using the skimr package to inspect the cleaned data
skim(cholera_linelist_clean)
```

# Export data to the "processed_data" folder using Rio() and Here()

In the same spirit that we *imported* raw files earlier (`rio::import(here("raw_data", …))`), we now *export* our cleaned line-list with two helper packages:

• **`rio`** – wraps dozens of I/O functions behind a single verb, `export()`. The file type is inferred from the extension you give it (`.csv`, `.xlsx`, `.rds`, etc.).\
• **`here`** – builds file paths that always start at the project root, so your code runs on any machine without fiddling with the working directory.

```{r}
# export the cleaned line-list to an Excel file in the "processed_data" folder
# packages used: rio, here, openxlsx
export(cholera_linelist_clean, here('processed_data', 'cholera_linelist_clean.xlsx'))
```

### Step 3: Group exercise

#### In groups of 4 or 5 do similar importing, inspecting, cleaning, and exporting for the dengue file in the raw_data folder

- Navigate to the student_code folder and open *Session 1: Tidyverse, importing data, data cleaning and validation_exercise.qmd* and complete the exercise
- You will be doing the same things we just covered on the *dengue_linelist_2025.xlsx* in the "raw_data" folder