---
title: "Session 2: Epidemiologic Measures, Tables, Plots & Maps"
author: "Max Wegener"
format: html
editor: visual
---

# Overview

This session extends the data-cleaning skills from Session 1 into basic outbreak epidemiology and spatial visualisation.\
Skills you will practise:

| Aim | Tool / function |
|-------------------|-----------------------------------------------------|
| 1\. Produce clean descriptive tables | **janitor::tabyl()**, **gtsummary**, **dplyr::count()** |
| 2\. Compute measures of association | 2×2 tables ➜ **epitools::oddsratio()** & **epiR::epi.2by2()** |
| 3\. Calculate attack rates | **dplyr** pipelines |
| 4\. Draw an epidemic curve | **ggplot2** |
| 5\. Map state-level incidence | **sf**, **ggplot2**, **scale_fill_viridis_c** |

All code is fully reproducible: press **Render** to rebuild the HTML after any change.

# Packages used for this session

```{r}
pacman::p_load(
  here,        # file paths
  rio,         # import/export
  tidyverse,   # core data wrangling + ggplot2
  janitor,     # tabyl()
  skimr,       # quick inspection
  gtsummary,   # publication-ready tables
  epitools,    # odds ratio
  epiR,        # risk ratio, attack rate helpers
  lubridate,   # date handling
  sf,          # simple features for GIS
  viridis,     # colour scales
  gt,
  janitor,
  epitools,
  rstatix,
  rio,          # file import/export
  here,         # relative filepaths 
  lubridate,    # working with dates/epiweeks
  aweek,        # alternative package for working with dates/epiweeks
  incidence2,   # epicurves of linelist data
  i2extras,     # supplement to incidence2
  stringr,      # search and manipulate character strings
  forcats,      # working with factors
  RColorBrewer, # Color palettes from colorbrewer2.org
  tidyverse     # data management + ggplot2 graphics
)
```

# Load the cleaned cholera line-list

```{r}
cholera_linelist <- import(here("processed_data", "cholera_linelist_clean.xlsx"))
```

```{r}
cholera_controls <- import(here("processed_data", "cholera_controls.xlsx"))
```

```{r}
cholera_cases_controls <- bind_rows(cholera_linelist, controls) %>% 
  mutate(case_status = coalesce(case_status, "case"))
```

# Descriptive tables

## Simple frequency table

```{r}
cholera_linelist %>% 
  tabyl(state_name, show_missing_levels = TRUE) %>% 
  adorn_pct_formatting()
```

```{r}
cholera_linelist %>% 
  tabyl(sex, outcome)
```

```{r}
cholera_linelist %>% 
  tabyl(sex, outcome) %>% 
        adorn_totals("row") %>%        # optional Total row
        adorn_percentages("row") %>% 
        adorn_pct_formatting(digits = 1) %>% 
        adorn_ns(position = "front") 
```

# Simple statistical tests

## t-test

```{r}
t.test(age ~ sex, data = cholera_linelist)
```

```{r}
cholera_linelist %>% 
  select(age, sex) %>%             # keep variables of interest
  tbl_summary(                               # produce summary table
    statistic = age ~ "{mean} ({sd})", # specify what statistics to show
    by = sex) %>%                        # specify the grouping variable
  add_p(age ~ "t.test") 
```

## Chi-squared test

```{r}
cholera_linelist %>% 
  tabyl(sex, outcome) %>% 
  select(-1) %>% 
  chisq_test()
```

```{r}
cholera_linelist %>% 
  select(sex, outcome) %>%    # keep variables of interest
  tbl_summary(by = outcome) %>%  # produce summary table and specify grouping variable
  add_p()  
```

## Summary statistics

```{r}
# ── packages ─────────────────────────────────────────────
library(dplyr)
library(gtsummary)
library(gt)            # {gt} comes with {gtsummary} for pretty output

# ── designate variable types ─────────────────────────────
cont_vars <- c("age")

cat_vars  <- c(
  "age_group", "state_name", "sex", "dehydration", "treatment", "water_source",
  "watery_diarrhea", "vomiting", "leg_cramps",
  "dehydration_signs", "fever"
)

# ── build Table 1: stratify by outcome (recovered vs death) ─────────────
tbl1 <- cholera_linelist %>% 
  select(outcome, all_of(cont_vars), all_of(cat_vars)) %>% 
  mutate(
    across(all_of(cat_vars), as.factor),     # be sure categoricals are factors
    outcome = factor(outcome,
                     levels = c("recovered", "death", "hospitalised"))   # set column order
  ) %>% 
  tbl_summary(
    by        = outcome,
    type      = list(all_of(cont_vars) ~ "continuous"),
    statistic = list(
                 all_continuous()  ~ "{mean} ± {sd}",     # change to median/IQR if desired
                 all_categorical() ~ "{n} ({p}%)"
               ),
    missing   = "no"                                     # omit explicit missing rows
  ) %>% 
  add_n()                             %>%  # overall N per column
  add_p(test = list(                     # χ² for cat, t-test for cont
           all_categorical() ~ "chisq.test",
           all_continuous()  ~ "t.test"
         )) %>% 
  bold_labels()                      %>%  # bold variable names
  modify_spanning_header(all_stat_cols() ~ "**Outcome**") %>%
  modify_header(label = "**Variable**")   %>%
  modify_footnote(
    p = "Categorical: χ² test; Continuous: two-sample t-test."
  )

# ── display in notebook / Jupyter / Quarto ───────────────
tbl1
```

## odds ratio

## risk ratio

## attack rate

# Epidemic Curves

```{r}# check range of onset dates}
ggplot(data = cholera_linelist)+
  geom_histogram(aes(x = onset_date))
```

# Maps

# Measures of association

Assume we want to know whether drinking surface water (vs. other sources) is associated with death.