---
title: "Intro to Positron and R Programing"
author: "Sofia Fertuzinhos"
format: html
editor: visual
---

# Location. Location. Location: Setting up the Working Directory**

## What is a Working Directory?**

The working directory is the folder on your computer where R will look for files (e.g., data) and save outputs by default. Think of it as telling R: \*"This is the folder we are working in today."\*

------------------------------------------------------------------------

### Step 1: Check Your Current Working Directory**

Use \`getwd()\` to see where R is currently pointed to on your computer. It will return a file path in the console, for example:

``` illustration
[1] "/Users/yourname/Documents"
```

```{r}
#| label: getwd
#| eval: false
getwd()
```

\> 💡 **\*\*Tip:\*\*** To learn more about **\*\*any\*\*** function in R, type \`?\` followed \> by its name in the console and press Enter. Try it: \> \`\`\` \> ?getwd \> ?setwd \> \`\`\`

------------------------------------------------------------------------

### Step 2: Set Your Working Directory**

We want R to point to the folder where **\*\*your data is stored\*\***. There are two ways to do this:

#### ✅ Option A — Point-and-Click \*(recommended for beginners)\***

[1.]{.underline} Look at the **\*\*bottom-right panel\*\*** in RStudio and click the **\*\*Files\*\*** tab.

[2.]{.underline} Navigate to the folder where your data is saved.

[3.]{.underline} Click the **\*\*gear icon (⚙)\*\*** in the Files toolbar.

[4.]{.underline} Select **\*\*"Set As Working Directory"\*\***.

\> 👀 Watch the Console — Positron will automatically run a \`setwd()\` command \> for you, and the path will be printed there.

#### ✅ Option B — Programmatic**

Use \`setwd()\` and provide the full path to your data folder. Replace the example path below with **\*\*your\*\*** folder's actual path:

```{r}
#| label: setwd
#| eval: false
# setwd("~/Documents/my_data_folder") # <-- edit this path and remove # before runing cell
```

To confirm it worked, run \`getwd()\` again:

```{r}
#| eval: false 
getwd()
```

\`\`\` \`\`\`

------------------------------------------------------------------------

### ⚠ Common Pitfalls**

::: callout-warning
**\*\*Windows users:\*\*** R uses forward slashes (\`/\`) or double backslashes (\`\\\\\`), **\*\*not\*\*** single backslashes (\`\\\`) like Windows Explorer shows.

❌ Wrong:

setwd("C:\\Users\\yourname\\Documents\\data") \`\`\`

✅ Right:

setwd("C:/Users/yourname/Documents/data")

\# or

setwd("C:\\\\Users\\\\yourname\\\\Documents\\\\data")
:::

::: callout-warning
The path must point to a **\*\*folder\*\***, not a file.
:::

::: callout-warning
\`setwd()\` only lasts for your **\*\*current R session\*\***. Each time you restart R, you will need to set it again — unless you use an Positron Project \*(introduced later in this course)\*.
:::

------------------------------------------------------------------------

### 🚀 A Better Long-Term Habit: RStudio Projects**

As you grow more comfortable with R, consider using **\*\*Positron Projects\*\*** (\`.Rproj\`). They automatically set the working directory to your project folder every time you open them — no \`setwd()\` needed! We will cover RStudio Projects later in this workshop.