---
title: "Session 3: Advanced Epidemic Curves & Spatial Mapping_Exercise"
author: "Max Wegener"
format: html
editor: visual
---

# Load packages

```{r}
pacman::p_load(
  here,
  rio,
  tidyverse,
  janitor,
  lubridate,
  sf,
  scales,
  plotly,
  slider,
  RColorBrewer
)
```

# Exercise

**Use the code shown during the instruction for the cholera data as a guide. The coding pipeline and logic are identical—you just need to apply them to the Dengue dataset and shapefiles!**

## Step 1: Load and Prepare the Cleaned Data

Import the cleaned Dengue dataset you exported during Session 1 (dengue_linelist_clean.xlsx) from the processed_data folder. Call the object dengue_linelist. Ensure you explicitly verify or coerce the onset_date column to an R Date format, and ensure your geospatial coordinate columns are numeric types.

```{r}
# Import cleaned dengue line-list, verify dates, and check coordinates
dengue_linelist <- import(here("processed_data", "dengue_linelist_clean.xlsx")) %>%
  mutate(
    onset_date = as.Date(onset_date),
    longitude  = as.numeric(longitude),
    latitude   = as.numeric(latitude)
  )
```

## Step 2: Constructing a Baseline Daily Epidemic Curve

Using ggplot2 and geom_histogram(), construct a daily epidemic curve tracking the progression of this Dengue outbreak.

- Set the bars to represent daily increments (binwidth = 1).

- Use a clear, distinct border and fill color for the bars.

```{r}
# Generate the baseline daily epidemic curve
# HINT: See Basic Epidemic Curve (Daily Intervals) on instructor file

```

## Step 3: Stratified Weekly Facet Analysis

Daily charts can sometimes mask wide-scale transmission velocities. Let's aggregate our temporal records into standard 7-day blocks and review regional patterns.

- Set the histogram to pool data into weekly increments.

- Map the Dengue virus serotype variable to the fill aesthetic.

- Use facet_wrap() to generate separate timelines for each state_name.

```{r}
# Generate the facetted weekly epidemic curve stratified by Dengue serotype
# HINT: See Facetted & Segmented Outbreak Overviews on instructor file


```

## Step 4: Localized Deep Dive with a 7-Day Trailing Moving Average

Identify the highest-burden state from your facetted plot. We will construct an advanced, publication-ready epidemic curve for that state to isolate operational transmission cycles from weekend reporting drops.

**Hint: Find the placeholders (.A. and .B. and .C. and .D.) in the code chunk below and replace them with the appropriate items:**

- .A. Dataframe name

- .B. The specific high-burden state string value to filter by (e.g., "Red Sea")

- .C. Date of symptom onset variable name

- .D. Categorical variable name tracking clinical status (e.g., "outcome")

```{r}
# 1. Prepare data aggregated by epidemiological week for the selected state
dengue_weekly_trends <- .A. %>% 
  filter(state_name == ".B." & !is.na(3)) %>% 
  mutate(epi_week = floor_date(.C., unit = "week", week_start = 1)) %>% 
  group_by(epi_week, .D.) %>% 
  summarise(weekly_cases = n(), .groups = "drop") %>% 
  
  # Calculate total cases per week to feed the rolling average line
  group_by(epi_week) %>% 
  mutate(
    total_weekly_cases = sum(weekly_cases),
    avg_7day_rolling = slider::slide_index_dbl(
      total_weekly_cases,
      .i = epi_week,
      .f = ~mean(.x, na.rm = TRUE),
      .before = 2  # Current week plus 2 weeks prior
    )
  ) %>% 
  ungroup()

# 2. Plot the weekly epicurve with the rolling average trendline
ggplot(data = dengue_weekly_trends) +
  geom_bar(
    mapping = aes(x = epi_week, y = weekly_cases, fill = outcome),
    stat = "identity", position = "stack", color = "white", width = 6, alpha = 0.8
  ) +
  geom_line(
    mapping = aes(x = epi_week, y = avg_7day_rolling, lty = "7-Week Rolling Trend"),
    color = "black", linewidth = 1.2
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_date(date_breaks = "2 weeks", labels = label_date_short(), expand = c(0.05, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  labs(
    title = "Weekly Dengue Outbreak Velocity & Trajectory Matrix",
    x = "Epidemiological Week (Start Date)", y = "Cases per Week (n)",
    fill = "Patient Outcome", lty = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Step 5: Denominator-Standardized Mapping

To target vector-control resources effectively, we must calculate population-adjusted incidence rates rather than mapping raw case counts (avoiding the "Choropleth Trap").

- Load the provided sudan_adm1 spatial object.

- Recode unmatched string characters (e.g., "Southern Darfur" to "South Darfur").

- Join the 18-state population matrix to establish proper risk denominators.

**Hint: Find the placeholders (.A. and .B. and .C.) in the code chunk below and replace them with the appropriate items:**

- .A. The original shapefile object name containing administrative boundaries

- .B. The population reference dataframe containing baseline denominators

- .C. The specific mathematical variable calculation for population standardization

```{r}
# 1. Establish validated national population matrix for Sudan (All 18 States)
state_pops <- tibble(
  state_name = c(
    "Khartoum", "South Darfur", "Gezira", "North Kordofan", 
    "Kassala", "Central Darfur", "White Nile", "North Darfur", 
    "Al Qadarif", "South Kordofan", "Sennar", "West Darfur", 
    "River Nile", "Red Sea", "West Kordofan", "East Darfur", 
    "Blue Nile", "Northern"
  ),
  population = c(
    7993900, 5353025, 5096920, 3174029, 
    2519071, 2499000, 2493880, 2304950, 
    2208385, 2107623, 1918692, 1775945, 
    1511442, 1482053, 1724897, 1587200, 
    1108391,  936255
  )
)

# 2. Harmonize Shapefile Strings on the fly
sudan_adm1_clean <- .A. %>%
  mutate(
    state_name = case_when(
      woe_name == "Southern Darfur" ~ "South Darfur",
      woe_name == "North Kordufan"  ~ "North Kordofan",
      woe_name == "Gedarif"         ~ "Al Qadarif",
      woe_name == "Eastern Darfur"  ~ "East Darfur",
      woe_name == "Western Darfur"  ~ "West Darfur",
      TRUE                          ~ woe_name
    )
  ) %>%
  group_by(state_name) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

# 3. Aggregate Dengue Line List Counts 
dengue_counts <- dengue_linelist %>%
  mutate(
    state_name = if_else(state_name %in% state_pops$state_name, state_name, "Other/Unknown")
  ) %>%
  count(state_name, name = "n_cases")

# 4. Execute Left-Join and compute population-standardized risk indices
dengue_map_df <- sudan_adm1_clean %>%
  left_join(dengue_counts, by = "state_name") %>%
  left_join(.B., by = "state_name") %>%
  mutate(
    n_cases = replace_na(n_cases, 0),
    # Calculate stable incidence rates per 100,000 residents
    cases_per_100k = (.C.)
  )

# 5. Plot the final standardized risk map
ggplot(dengue_map_df) +
  geom_sf(aes(fill = cases_per_100k), colour = "grey40", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", direction = -1, na.value = "grey70") +
  labs(
    title = "National Dengue Incidence Matrix, Sudan 2025",
    subtitle = "Population-Standardized Attack Rates (Denominator Adjusted)",
    fill  = "Cases per 100k"
  ) +
  theme_minimal()
```

## \## Step 6: Interactive Dashboard Point Map Component

Bypass political boundaries by plotting precise coordinate intersections. Convert a static point distribution into a dynamic, hover-responsive map tracking local vector controls.

**Hint: Find the placeholders (.A., .B., .C., .D., and .E.) in the code chunk below and replace them with the appropriate items:**

- **.A.** Call the correct dataframe we have been working with in this exercise.

- **.B.** Add the longitude variable name.

- **.C.** Add the latitude variable name.

- **.D.** Provide the name of the new spatial dataset (`sf` object) we created in part 1.

- **.E.** Add the categorical variable name we want to color-code our points by (e.g., tracking patient status).

```{r}
# 1. Extract valid points and assign spatial coordinates
dengue_pts <- .A. %>%
  filter(!is.na(.B.) & !is.na(.C.)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

sudan_boundaries <- st_transform(sudan_adm1, 4326)

# 2. Build the static ggplot with custom hover labels
dengue_point_map <- ggplot() +
  geom_sf(data = sudan_boundaries, fill = "grey95", colour = "grey85", linewidth = 0.2) +
  geom_sf(
    data = .D.,
    aes(
      colour = .E.,
      text = paste("Age:", age, "<br>Outcome:", outcome, "<br>Netting:", mosquito_net)
    ),
    alpha = 0.6, size = 1.5
  ) +
  coord_sf(xlim = c(21, 39), ylim = c(9, 22)) +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "Interactive Dengue Field Surveillance Points", colour = "Patient Outcome") +
  theme_minimal()

# 3. Convert to an interactive plotly widget
ggplotly(dengue_point_map, tooltip = "text")
```