---
title: "Session 1: Tidyverse, importing data, data cleaning and validation_Exercise"
author: "Max Wegener"
format: html
editor: visual
---

# Exercise

- Select one computer to perform the exercise

- To show that you have completed the exercise, you group will share code in the file as well as the state_name counts output and the skim() output for the cleaned dataset.

**HINT: Use the code shown during the instruction for eh cholera data as a guide, the coding and flow are pretty much identical**

## Step 1: Importing Data

Import *dengue_linelist_2025_sudan.xlsx* using the here() and rio() packages and call the object "dengue_linelist_raw"

```{r}

```

## Step 2: Data Cleaning and Validation

- Start by upodaitng the name of the DF to "dengue_linelist_clean*"*

- Clean the column names

- Convert date columns from POSIXct to an R Date

- Rename columns to be more descriptive

  | Old Name     | New Name      |
  |--------------|---------------|
  | `date_onset` | `onset_date`  |
  | `lat`        | `latitiude`   |
  | `lon`        | `longitutde`  |
  | `days_hosp`  | hospital_days |

- Standardize messy state name values into clean, consistent categories

- Change blanks or NA values to "unknown"

- Select only the variables we need for analysis and arrange them

```{r}
#add cleaning code chunk here


```

Here is where you will count the new state_name variable to ensure you address all the spelling mistakes

```{r}
# count state_name here
# code from instruction: cholera_linelist_clean %>% count(state_name)

```

Here, you will use skim() to make sure all the NAs and blanks are addressed

```{r}
# use skim() to reinspect the cleaned data
```

## Step 3: Exporting Cleaned Data

Using here() and rio(), export the cleaned dataset to the "processed_data" folder and call the new file "dengue_linelist_clean.xlsx"

```{r}
# export the cleaned line-list to an Excel file in the "processed_data" folder


```