---
title: "Session 2: Summaries, Measures, Tables, & Epi-curves_Exercise"
author: "Max Wegener"
format: html
editor: visual
---

# Load packages

```{r}
pacman::p_load(
  here,
  rio,
  tidyverse,
  janitor,
  gtsummary,
  rstatix
)
```

# Exercise

- Select one computer to perform the exercise.

- To show that you have completed the exercise, your group will share your code (with demonstration to see console), the final publication-ready "Table 1" output, and your facetted epidemic curve plot.

**HINT: Use the code shown during the instruction for the cholera data as a guide. The coding pipeline and logic are identical—you just need to apply them to the Dengue dataset!**

------------------------------------------------------------------------

## Step 1: Load the Cleaned Data

Import the cleaned Dengue dataset you exported during Session 1 (*dengue_linelist_clean.xlsx*) from the `processed_data` folder using `here()` and `rio()`. Call the object `dengue_linelist`. Ensure you explicitly verify or coerce the `onset_date` column to an R Date format.

```{r}
# Import cleaned dengue line-list and verify the onset_date format

dengue_linelist <- import(here("processed_data", "dengue_linelist_clean.xlsx")) %>%
  mutate(onset_date = as.Date(onset_date))
```

## Step 2: Descriptive Frequency Tables

Using the {janitor} package, generate descriptive tables to understand your study population. Create a simple frequency table showing the distribution of cases across different state_name values, including formatted percentages.

```{r}
# 1. Simple frequency table for state_names
# HINT: See Simple frequency table on instructor file


```

Create a two-way cross-tabulation looking at sex against clinical outcome. Adorn the table to show row totals and row percentages front-facing (N (Pct%)).

```{r}
# 2. Adorned two-way table for sex vs. outcome
# HINT: see Advanced multi-adorned table on instructor file


```

## Step 3: Summary Statistics & Simple Hypothesis Testing

We want to evaluate the demographic profile of our cases.

- Use {rstatix} to calculate the overall summary statistics for age.
- Perform a statistical hypothesis test to determine if the continuous variable age differs significantly across different clinical outcome groups (e.g., recovered vs. death vs. hospitalised).

**Hint: Check the distribution of age first to decide if you should use a parametric (ANOVA) or non-parametric (Kruskal-Wallis) approach!**

```{r}
#HINT: see Overall summary stats for age on instructor file


```

For this let's compare the age distribution of dengue cases between different outcomes

```{r}
# Run the appropriate hypothesis test for age across outcome groups

# HINT: Use non-parametric test used to compare a contiuous varaible to a variable with more than two values



```

## Step 4: Constructing a Publication-Ready "Table 1"

Using the {gtsummary} package, assemble a comprehensive baseline characteristics table (tbl_summary) stratified by outcome.

- Include age, sex, state name, and key symptoms (like fever or joint pain).
- Add an overall sample size column (add_n()).
- Add p-values (add_p()) using appropriate tests for continuous and categorical variables.
- Bold the labels and add a professional spanning header called "Clinical Outcome".

**Hint: Find the following numbers in the code chunk and replace with the appropriate words. See instructor code.**

Any number within quotations or asterisks (\*) the appropriate names/labels should be between them as well

1.  Dataframe name
2.  Variable name
3.  test name
4.  test name
5.  Lable name for spanning header
6.  Lable name for header

```{r}

table_one <- 1 %>% 
  select(outcome, age, sex, state_name, serotype, fever, severity, retroorbital_pain) %>% 
  mutate(across(c(sex, state_name, fever, retroorbital_pain), as.factor)) %>% 
  tbl_summary(
    by = 2,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no"
  ) %>% 
  add_n() %>% 
  add_p(
    test = list(
      all_categorical() ~ "3", 
      all_continuous() ~ "4"
    )
  ) %>% 
  bold_labels() %>% 
  modify_spanning_header(all_stat_cols() ~ "** 5 **") %>% 
  modify_header(label = "** 6 **")

table_one
```

## Step 5: Draw a Facetted Epidemic Curve

Using ggplot2, build an epidemic curve to visualize the temporal progression of the Dengue outbreak. Map onset_date to the X-axis using geom_histogram().

- Group the cases into 7-day epidemiological weeks by adjusting the binwidth.
- Color the bars dynamically by setting the fill parameter to track case counts by water_source or outcome.
- Use facet_wrap() to split the epidemic curves into parallel timelines for each individual state_name.
- Clean up the presentation with theme_minimal() and add descriptive titles and axis labels.

**Hint: Find the following numbers in the code chunk and replace with the appropriate words. See instructor code.**

Any number within quotations, the appropriate names/labels should be between the quotations

1.  dataframe name
2.  the data for the x-axis
3.  stratifying variable
4.  x-axis label
5.  y-axis label
6.  fill label

Notice that "binwidth" has a question (?) - replace with the number of days the bin should be

**Hint: Again, see instructor file**

```{r}
# Create your facetted weekly epi-curve here
# Create your facetted weekly epi-curve here
ggplot(data = 1, mapping = aes(x = 2, fill = 3)) +
  geom_histogram(
    binwidth = ? ,             # Groups dates into 7-day epi-weeks
    position = "stack",       # Stacks outcomes on top of each other
    color = "white",          # Adds clean borders between bars
    linewidth = 0.2
  ) +
  scale_fill_brewer(palette = "Dark2") +
  #scale_fill_viridis_d(option = "rocket") +
  #scale_fill_viridis_d(option = "plasma") +
  facet_wrap(~ state_name, scales = "free_y") + # Allows y-axes to scale to individual state sizing
  theme_minimal() +
  labs(
    title = "Weekly Incidence of Dengue Cases across Sudan States (2025)",
    subtitle = "Aggregated by 7-day epidemiological windows",
    x = "4",
    y = "5",
    fill = "6"
  ) +
  theme(
    strip.text = element_text(face = "bold"), # Bolds the facet state names
    legend.position = "bottom"
  )
```