---
title: "Session 1: Tidyverse, importing data, data cleaning and validation_Exercise"
author: "Max Wegener"
format: html
editor: visual
---

# Load packages

```{r}
pacman::p_load(
  here,  
  rio,         
  tidyverse,
  janitor,    
  lubridate,    
  matchmaker,   
  epikit,    
  skimr       
)
```

# Exercise

**Use the code shown during the instruction for eh cholera data as a guide, the coding and flow are pretty much identical**

## Step 1: Importing Data

Import *dengue_linelist_2025_sudan.xlsx* using the here() and rio() packages and call the object "dengue_linelist_raw"

**Reminder!** we are importing from the *raw_data* folder

```{r}
# Finish writing the code to import the data here by repalcing each "..."
dengue_linelist_raw <- improt(here("...", "..."))
```

## Step 2: Data Cleaning using the tidyverse

Using the Tidyverse packages, we build one single cleaning pipeline using the pipe operator (`%>%`). Each pipe passes the result of one step to the next.

**Hint: Find the following numbers in the code chunk and replace with the appropriate words. See instructor code.**

Any number within quotations or asterisks (\*) the appropriate names/labels should be between them as well

1.  New Dataframe name
2.  Function name Variable name
3.  New column name (see table below)
4.  New column name (see table below)
5.  New column name (see table below)
6.  New column name (see table below)
7.  New column name for state variable
8.  current varaible name for male and female values
9.  New column name for age group variable
10. New value to replace blanks and NAs in character columns

-   Start by updating the name of the DF to "dengue_linelist_clean*"*

-   Clean the column names

-   Convert date columns from POSIXct to an R Date

-   Rename columns to be more descriptive

    | Old Name     | New Name        |
    |--------------|-----------------|
    | `date_onset` | `onset_date`    |
    | `lat`        | `latitiude`     |
    | `lon`        | `longitutde`    |
    | `days_hosp`  | `hospital_days` |

-   Standardize messy state name values into clean, consistent categories

-   Change blanks or NA values to "unknown"

-   Select only the variables we need for analysis and arrange them

```{r}
#add cleaning code chunk here
1 <- dengue_linelist_raw %>% 
  # STEP 2a: Clean column names to be consistent and easier to work with
  2() %>%
  # STEP 2b: Convert date columns from character strings to proper R Date objects
  mutate(date_onset = as.Date(date_onset)) %>%
  # STEP 2c: Rename columns to be more descriptive and consistent
  rename(
    3   = date_onset,
    4     = lat,
    5   = lon,
    6 = days_hosp
  ) %>%
 mutate(
    # ----- State -------------------------------------------------------
    7 = case_when(
      str_detect(state, regex("blu|blue",  ignore_case = TRUE)) ~ "Blue Nile",
      str_detect(state, regex("gez",       ignore_case = TRUE)) ~ "Gezira",
      str_detect(state, regex("kass",      ignore_case = TRUE)) ~ "Kassala",
      str_detect(state, regex("^khar",     ignore_case = TRUE)) ~ "Khartoum",
      str_detect(state, regex("north",     ignore_case = TRUE)) ~ "North Darfur",
      str_detect(state, regex("red",       ignore_case = TRUE)) ~ "Red Sea",
      str_detect(state, regex("south",     ignore_case = TRUE)) ~ "South Darfur",
      str_detect(state, regex("white",     ignore_case = TRUE)) ~ "White Nile",
      is.na(state) | str_trim(state) == ""                       ~ "Other/Unknown",
      TRUE                                                       ~ "Other/Unknown"
    ),
    
    # ----- Sex ---------------------------------------------------------
    8 = case_when(
      str_detect(sex, regex("^m$",      ignore_case = TRUE)) ~ "male",
      str_detect(sex, regex("^male$",   ignore_case = TRUE)) ~ "male",
      str_detect(sex, regex("^f$",      ignore_case = TRUE)) ~ "female",
      str_detect(sex, regex("^female$", ignore_case = TRUE)) ~ "female",
      is.na(sex) | str_trim(sex) == ""                       ~ "unknown",
      TRUE                                                   ~ "unknown"
    )
  ) %>% 
  
  # STEP 2e ── ordered age-group factor ---------------------------
  mutate(
    9 = cut(
      age,
      breaks = c(0, 5, 15, 30, 45, 60, Inf),  
      right   = FALSE,                         
      include.lowest = TRUE,
      labels = c("0–4", "5–14", "15–29",
                 "30–44", "45–59", "60+")
    )
  ) %>%
  # STEP 2f: Change blanks or NA values to "unknown" 
  mutate(
    across(
      where(is.character),
      ~ replace_na( na_if(trimws(.x), ""), "10" )
    )
  ) %>%
  select(case_id, onset_date, state_name, age_group, everything(), -state)

```

## Step 3: Reinspect the cleaned dataframe

Here is where you will count the new state_name variable to ensure you address all the spelling mistakes

```{r}
# Hint: DF_NAME %>% count(VARIABLE_NAME) to count the state_name variable
# count state_name here


```

Here, you will use skim() to make sure all the NAs and blanks are addressed

```{r}
# HINT: Use skim() to re-inspect the newly cleaned dengue dataframe

```

## Step 4: Exporting Cleaned Data

Using here() and rio(), export the cleaned dataset to the "processed_data" folder and call the new file "dengue_linelist_clean.xlsx"

```{r}
# export the cleaned line-list to an Excel file in the "processed_data" folder


```