---
title: "Session 2: Summaries, Measures, Tables, & Epi-curves_Exercise"
author: "Max Wegener"
format: html
editor: visual
---

# Load packages

```{r}
pacman::p_load(
  here,
  rio,
  tidyverse,
  janitor,
  gtsummary,
  rstatix
)
```

# Exercise

**Use the code shown during the instruction for the cholera data as a guide. The coding pipeline and logic are identical—you just need to apply them to the Dengue dataset!**

------------------------------------------------------------------------

## Step 1: Load the Cleaned Data

Import the cleaned Dengue dataset you exported during the Session 1 exercise (*dengue_linelist_clean.xlsx*) from the `processed_data` folder using `here()` and `rio()`. Call the object `dengue_linelist`. Ensure you explicitly verify or coerce the `onset_date` column to an R Date format.

```{r}
# Import cleaned dengue line-list and verify the onset_date format

dengue_linelist <- import(here("processed_data", "dengue_linelist_clean.xlsx")) %>%
  mutate(onset_date = as.Date(onset_date))
```

## Step 2: Descriptive Frequency Tables

Using the `janitor` package's `tabyl()` function, generate descriptive tables to understand your study population. Create a simple frequency table showing the distribution of cases across different state_name values, including formatted percentages using the `adorn_pct_formatting()` function.

```{r}
# 1. Simple frequency table for state_names
# HINT: See Simple frequency table on instructor file


```

# Load packages

```{r}
pacman::p_load(
  here,
  rio,
  tidyverse,
  janitor,
  gtsummary,
  rstatix
)
```

# Exercise

**Use the code shown during the instruction for the cholera data as a guide. The coding pipeline and logic are identical—you just need to apply them to the Dengue dataset!**

------------------------------------------------------------------------

## Step 1: Load the Cleaned Data

Import the cleaned Dengue dataset you exported during the Session 1 exercise (*dengue_linelist_clean.xlsx*) from the `processed_data` folder using `here()` and `rio()`. Call the object `dengue_linelist`. Ensure you explicitly verify or coerce the `onset_date` column to an R Date format.

```{r}
# Import cleaned dengue line-list and verify the onset_date format

dengue_linelist <- import(here("processed_data", "dengue_linelist_clean.xlsx")) %>%
  mutate(onset_date = as.Date(onset_date))
```

## Step 2: Descriptive Frequency Tables

Using the `janitor` package's `tabyl()` function, generate descriptive tables to understand your study population. Create a simple frequency table showing the distribution of cases across different state_name values, including formatted percentages using the `adorn_pct_formatting()` function.

```{r}
# 1. Simple frequency table for state_names
# HINT: See Simple frequency table on instructor file

dengue_linelist %>% 
  tabyl(state_name) %>% 
  adorn_pct_formatting()
```

Create a two-way cross-tabulation looking at `sex` and `outcome`.

Adorn the table to show **row totals** and front-facing row percentages with a single decimal point.

```{r}
# 2. Adorned two-way table for sex vs. outcome
# HINT: see Advanced multi-adorned table on instructor file


```

## Step 3: Summary Statistics & Simple Hypothesis Testing

We want to evaluate the demographic profile of our cases.

- Use the `rstatix` package to calculate the overall summary statistics for age.
- Perform a statistical hypothesis test to determine if the continuous variable **age** differs significantly across different clinical **outcome** groups (e.g., recovered vs. death vs. hospitalised).

**Hint: Check the distribution of age first to decide if you should use a parametric (ANOVA) or non-parametric (Kruskal-Wallis) approach!**

```{r}
#HINT: see Overall summary stats for age on instructor file


```

Now, let's compare the age distribution of dengue cases between different **outcomes**

```{r}
# Run the appropriate hypothesis test for age across outcome groups

# HINT: Use non-parametric test used to compare a contiuous varaible to a variable with more than two values



```

## Step 4: Constructing a Publication-Ready "Table 1"

Using the `gtsummary` package, assemble a comprehensive baseline characteristics table `tbl_summary` stratified by **outcome**.

- Include age, sex, state name, and key symptoms (like fever or joint pain).
- Add an overall sample size column (add_n()).
- Add p-values (add_p()) using appropriate tests for continuous and categorical variables.
- Bold the labels and add a professional spanning header called "Clinical Outcome".

**Hint: Find the following numbers in the code chunk and replace with the appropriate words. See instructor code.**

Any number within quotations or asterisks (\*) the appropriate names/labels should be between them as well

1.  Dataframe name
2.  Variable name
3.  test name
4.  test name
5.  Lable name for spanning header
6.  Lable name for header

```{r}

table_one <- 1 %>% 
  select(outcome, age, sex, state_name, serotype, fever, severity, retroorbital_pain) %>% 
  mutate(across(c(sex, state_name, fever, retroorbital_pain), as.factor)) %>% 
  tbl_summary(
    by = 2,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no"
  ) %>% 
  add_n() %>% 
  add_p(
    test = list(
      all_categorical() ~ "3", 
      all_continuous() ~ "4"
    )
  ) %>% 
  bold_labels() %>% 
  modify_spanning_header(all_stat_cols() ~ "** 5 **") %>% 
  modify_header(label = "** 6 **")

table_one
```

# Step 5: Case Fatality Risk / Ratio (CFR)

Calculate the clinical severity of this outbreak using your case-only surveillance data.

- First, write a pipeline using summarise() to calculate the overall global CFR for the entire dataset.

- Next, modify that pipeline using group_by() to calculate and compare the CFR across different states (**state_name**). Identify which state is experiencing the highest clinical lethality.

**Hint: Find the following numbers in the code chunk and replace them with the appropriate words or symbols. Dataframe name**

1.  The column name tracking the clinical outcome
2.  The specific character string value inside that column representing an individual who died (e.g., "dead" or "death")
3.  The variable you want to group by to look at regional differences

## Step 5.1 Calculate the global CFR for the entire dataset

```{r}
# Calculate global CFR for the entire dataset 
# Proportional Morbidity
global_cfr <- 1 %>% 
  summarise(
    cfr = sum(2 == "death") / n() * 100
  )

global_cfr
```

## Step 5.2 Calculate and compare the CFR across different states

Now, let's calculate a CRF for each state by adding a `group_by()` function above the `sumarise()` function and include **state_name.**

```{r}
# Calculate CFR by state
cfr_by_state <- 1 %>% 
  group_by(3) %>% 
  summarise(
    cfr = sum(2 == "death") / n() * 100
  ) %>% 
  arrange(desc(cfr)) # Sort states by CFR in descending order

cfr_by_state  
```

# Step 6: Proportional Morbidity (Exposure Proportions)

Because surveillance officers frequently operate without population denominators or control groups, you must evaluate shifts in exposure trends purely among known cases.

1.  Use the `dplyr` package's `filter()` and `between()` functions to extract cases occurring during the first week of the outbreak (e.g., January 1 to January 7, 2025).

2.  Pass this filtered data to the `janitor` packge's `tabyl()` function to look at the proportion of a key environmental risk factor like standing water proximity (water_container_status).

**Hint: Find the following numbers in the code chunk and replace them with the appropriate words or variable names.**

1.  Dataframe name
2.  Date of symptom onset variable
3.  Environmental risk factor/exposure variable name

## Step 6.1 Proportional Morbidity for Week 1 (Jan 1 - Jan 7, 2025)

```{r}
1 %>%
  filter(
    !is.na(2) & 
    between(2, as.Date("2025-01-01"), as.Date("2025-01-07"))
  ) %>%
  tabyl(3) %>%
  adorn_pct_formatting()
```

## Step 6.2 Proportional Morbidity for the Late Phase (e.g., Feb 1 - Feb 7, 2025)

Now we will adjust the date range to see what happens to the results and compare to the previous week (step 6.1)

**IMPORTANT: run the code the first time then comment out the first date line and uncomment the second date line to compare the two weeks**

```{r}
1 %>%
  filter(
    !is.na(2) & 
    between(2, as.Date("2025-02-01"), as.Date("2025-02-07"))
    #between(2, as.Date("2025-12-25"), as.Date("2025-12-31"))
  ) %>%
  tabyl(3) %>%
  adorn_pct_formatting()
```